---
title: "do calc"
author: "Simon Brewer"
date: "5/10/2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries and whatnot

From https://www.andrewheiss.com/blog/2021/09/07/do-calculus-backdoor

```{r warning=FALSE, message=FALSE}
library(tidyverse)  # For ggplot2 and friends
library(patchwork)  # For combining plots
library(ggdag)      # For making DAGs with ggplot
library(dagitty)    # For dealing with DAG math
library(latex2exp)  # Easily convert LaTeX into arcane plotmath expressions
library(ggtext)     # Use markdown in ggplot labels

# Create a cleaner serifed theme to use throughout
theme_do_calc <- function() {
  theme_dag(base_family = "Times New Roman") +
    theme(plot.title = element_text(size = rel(1.5)),
          plot.subtitle = element_markdown())
}

# Make all geom_dag_text() layers use these settings automatically
update_geom_defaults(ggdag:::GeomDagText, list(family = "Times New Roman", 
                                               fontface = "bold",
                                               color = "black"))
```

## Exploring the rules of do-calculus

Here’s what each rule actually does:

- Rule 1: Decide if we can ignore an observation
- Rule 2: Decide if we can treat an intervention as an observation
- Rule 3: Decide if we can ignore an intervention

Whoa! That's exceptionally logical. Each rule is designed to help simplify and reduce nodes in a DAG by either ignoring them (Rules 1 and 3) or making it so interventions like $\mbox{do}(\cdot)$ can be treated like observations instead (Rule 2).

Let's explore each of these rules in detail. In all these situations, we’re assuming that there’s a DAG with 4 nodes: $W$, $X$, $Y$, and $Z$. $Y$ is always the outcome; $X$ is always the main treatment. In each rule, our goal is to get rid of $Z$ by applying the rule. When talking about interventions in a graph, there’s a special notation with overlines and underlines:

- An overline like $G_{\overline{X}}$ means that you delete all the arrows going into $X$
- An underline like $G_{\underline{X}}$ means that you delete all the arrows coming out of $X$

I imagine this line like a wall:

- If the wall is on top of X like $\overline{X}$, you can’t draw any arrows going into it, so you delete anything going in
- If the wall is on the bottom of X like $\underline{X}$, you can’t draw any arrows going out of it, so you delete anything going out

### Rule 1: Ignoring observations

According to Rule 1, we can ignore any observational node if it doesn’t influence the outcome through any path, or if it is d-separated from the outcome. Here’s the formal definition:

\[
P(y|z,\mbox{do}(x)),w) = P(y|\mbox{do}(x),w) \quad \mbox{if} (Y \bot Z|W, X) G_{\overline{X}}
\]


There are a lot of moving parts here, but remember, the focus in this equation is 
$z$. Our goal here is to remove or ignore $z$. Notice how $z$ exists on the left-hand side of the equation and how it is gone on the right-hand side. As long as we meet the cryptic conditions of $(Y \bot Z|W, X) G_{\overline{X}}$, we can get rid of it. But what the heck does that even mean?

Here, $G_{\overline{X}$ means “the original causal graph with all arrows into $X$ removed”, while the $Y \bot Z|W, X$ part means "Y is independent of Z, given W and X" in the new modified graph. If the Y and Z nodes are d-separated from each other after we account for both W and X, we can get rid of Z and ignore it.

Let’s look at this graphically to help make better sense of this. We’ll use the `dagify()` function from **ggdag** to build a couple DAGs: one complete one ($G$) and one with all the arrows into X deleted ($G_{\overline{X}$). X causes both X and Y, while W confounds X, Y, and Z.

```{r}
rule1_g <- dagify(
  Y ~ X + W,
  X ~ W,
  Z ~ X + W,
  coords = list(x = c(X = 1, Y = 2, Z = 1.25, W = 1.75),
                y = c(X = 1, Y = 1, Z = 2, W = 2))
)

rule1_g_x_over <- dagify(
  Y ~ X + W,
  Z ~ X + W,
  coords = list(x = c(X = 1, Y = 2, Z = 1.25, W = 1.75),
                y = c(X = 1, Y = 1, Z = 2, W = 2))
) 
```

```{r}
plot_rule1_g <- ggplot(rule1_g, aes(x = x, y = y, 
                                    xend = xend, yend = yend)) +
  geom_dag_edges() +
  geom_dag_point(color = "grey80", size = 10) +
  geom_dag_text() +
  labs(title = TeX("$G$"),
       subtitle = "Original DAG") +
  theme_do_calc()

plot_rule1_g_x_over <- ggplot(rule1_g_x_over, aes(x = x, y = y, 
                                                  xend = xend, yend = yend)) +
  geom_dag_edges() +
  geom_dag_point(color = "grey80", size = 10) +
  geom_dag_text() +
  labs(title = TeX("$G_{\\bar{X}}$"),
       subtitle = "DAG with arrows *into* X deleted") +
  theme_do_calc()

plot_rule1_g | plot_rule1_g_x_over
```

If we want to calculate the causal effect of X on Y, do we need to worry about Z here, or can we ignore it? Let’s apply Rule 1. If we look at the modified $G_{\overline{X}}$, Y and Z are completely d-separated if we account for both W and X — there's no direct arrow between them, and there’s no active path connecting them through W or X, since we’re accounting for (or condition on) those nodes. Y and Z are thus d-separated and $Y \bot Z|W, X$. We can confirm this with the `impliedConditionalIndependencies()` function from the **dagitty** package:

```{r}
impliedConditionalIndependencies(rule1_g)
impliedConditionalIndependencies(rule1_g_x_over)
```

And there it is! The second independency there is $Y \bot Z|W, X$. That means that we can apply Rule 1 and ignore Z, meaning that

\[
P(y|z,\mbox{do}(x)),w) = P(y|\mbox{do}(x),w)
\]

This makes sense but is a little too complicated for me, since we’re working with four different nodes. We can simplify this and pretend that $\mbox{do}(x)$ is nothing and that X doesn’t exist. That leaves us with just three nodes — W, Y, and Z — and this DAG:

```{r}
rule1_g_simple <- dagify(
  Y ~ W,
  Z ~ W,
  coords = list(x = c(Y = 2, Z = 1, W = 1.5),
                y = c(Y = 1, Z = 1, W = 2))
)

plot_rule1_g_simple <- ggplot(rule1_g_simple, aes(x = x, y = y, 
                                                  xend = xend, yend = yend)) +
  geom_dag_edges() +
  geom_dag_point(color = "grey80", size = 10) +
  geom_dag_text() +
  labs(title = TeX("$G$"),
       subtitle = "Simplified DAG without X") +
  theme_do_calc()
plot_rule1_g_simple
```

The simplified X-free version of Rule 1 looks like this:

\[
P(y|z,w) = P(y|w) \quad \mbox{if} (Y \bot Z|W) G
\]

In other words, we can ignore Z and remove it from the $P(y|z,w)$ equation if Y and Z are d-separated (or independent of each other) after accounting for W. Once we account for W, there’s no possible connection between Y and Z, so they really are d-separated. We can again confirm this with code:

```{r}
impliedConditionalIndependencies(rule1_g_simple)
```

There we go. Because $Y \bot Z|W$, we can safely ignore Z

### Rule 2: Treating interventions as observations

Rule 1 is neat, but it has nothing to do with causal interventions or the 
$\mbox{do}(\cdot)$ operator. It feels more like a housekeeping rule — it’s a way of simplifying and removing unnecessary nodes that don’t have to do with the main treatment &larr; outcome relationship.

With Rule 2, we start messing with interventions. In an experiment like a randomized controlled trial, a researcher has the ability to assign treatment and either $\mbox{do}(x)$ or not $\mbox{do}(x)$. With observational data, though, it’s not possible to $\mbox{do}(x)$ directly. It would be fantastic if we could take an intervention like $\mbox{do}(x)$ and treat it like regular non-interventional observational data. Rule 2 lets us do this.

According to Rule 2, interventions (or $\mbox{do}(x)$) can be treated as observations (or $x$) when the causal effect of a variable on the outcome (X &rarr; Y) only influences the outcome through directed paths. The official math for this is this complicated thing:

\[
P(y | \mbox{do}(z),\mbox{do}(x),w) = P(y | z,\mbox{do}(x),w) \quad \mbox{if} (Y \bot Z|W, X) G_{\overline{X},\underline{Z}}
\]

For me, this is super confusing, since there are two different $\mbox{do}(\cdot)$ operators here and when I think of causal graphs, I think of single interventions. Like we did with Rule 1, we can simplify this and pretend that there’s no intervention $\mbox{do}(x)$ (we’ll do the full rule in a minute, don’t worry). Again, this is legal because each of these rules are focused on messing with the Z variable: ignoring it or treating it as an observation. That leaves us with this slightly simpler (though still cryptic) equation:

\[
P(y | \mbox{do}(z),w) = P(y | z,w) \quad \mbox{if} (Y \bot Z|W) G_\underline{Z}
\]

Notice how the left-hand side has the interventional $\mbox{do}(z)$, while the right-hand side has the observed $z$. As long as we meet the condition $(Y \bot Z|W) G_\underline{Z}$, we can transform $\mbox{do}(z)$ into $z$ and work only with observational data. Once again, though, what does this $(Y \bot Z|W) G_\underline{Z}$ condition even mean?

Here, $G_\underline{Z}$ means "the original causal graph with all arrows out of Z removed", while the $Y \bot Z|W$ part means "Y is independent of Z, given W" in the new modified graph. Similar to Rule 1, if the Y and Z nodes are d-separated from each other after we account for W, we can legally treat $\mbox{do}(z)$ like $z$.

As we did with Rule 1, we’ll build a couple basic DAGs: a complete one ($G$) and one with all the arrows out of Z deleted ($G_\underline{Z}$).

```{r}
rule2_g_simple <- dagify(
  Y ~ Z + W,
  Z ~ W,
  coords = list(x = c(Y = 2, Z = 1, W = 1.5),
                y = c(Y = 1, Z = 1, W = 2))
)

rule2_g_simple_z_under <- dagify(
  Y ~ W,
  Z ~ W,
  coords = list(x = c(Y = 2, Z = 1, W = 1.5),
                y = c(Y = 1, Z = 1, W = 2))
) 
```

```{r}
plot_rule2_g_simple <- ggplot(rule2_g_simple, 
                              aes(x = x, y = y, 
                                  xend = xend, yend = yend)) +
  geom_dag_edges() +
  geom_dag_point(color = "grey80", size = 10) +
  geom_dag_text() +
  labs(title = TeX("$G$"),
       subtitle = "Original DAG") +
  theme_do_calc()

plot_rule2_g_simple_z_under <- ggplot(rule2_g_simple_z_under, 
                                      aes(x = x, y = y, 
                                          xend = xend, yend = yend)) +
  geom_dag_edges() +
  geom_dag_point(color = "grey80", size = 10) +
  geom_dag_text() +
  labs(title = TeX("$G_{\\underline{Z}}$"),
       subtitle = "DAG with arrows *out of* Z deleted") +
  theme_do_calc()

plot_rule2_g_simple | plot_rule2_g_simple_z_under
```

So, can we treat Z here like an observational node instead of a interventional 
$\mbox{do}(z)$ node? Let’s apply Rule 2. If we look at the modified $G_\underline{Z}$ graph, Z and Y are completely d-separated if we account for W—there’s no direct arrow between them, and there’s no active path connecting them through W since we’re conditioning on W. We can thus say that $Y \bot Z|W$. We can confirm this with code too:

```{r}
impliedConditionalIndependencies(rule2_g_simple_z_under)
```

Woohoo! Because $Y \bot Z|W$ in that modified $G_\underline{Z}$ graph, we can legally convert the interventional $\mbox{do}(z)$ to just a regular old observational $z$:

\[
P(y | \mbox{do}(z),w) = P(y | z,w)
\]

So far we’ve applied Rule 2 to a simplified DAG with three nodes, but what does it look like if we’re using the full four-node graph that is used in the formal definition of Rule 2?

\[
P(y | \mbox{do}(z),\mbox{do}(x),w) = P(y | z,\mbox{do}(x),w) \quad \mbox{if} (Y \bot Z|W, X) G_{\overline{X},\underline{Z}}
\]

Here’s one graphical representation of a graph with the four nodes W, X, Y, and Z (but it’s definitely not the only possible graph! These do-calculus rules don’t assume any specific relationships between the nodes). Here, Y is caused by both X and Z, and we’ll pretend that they’re both interventions (so $\mbox{do}(x)$ and $\mbox{do}(z)$). X is causally linked to Z, and W confounds all three: X, Y, and Z. Graph $G$ shows the complete DAG; Graph ${\overline{X},\underline{Z}}$ shows a modified DAG with all arrows into X deleted ($\overline{X}$) and all arrows out of Z deleted ($\underline{Z}$).

```{r}
rule2_g <- dagify(
  Y ~ X + W + Z,
  X ~ W,
  Z ~ X + W,
  coords = list(x = c(X = 1, Y = 2, Z = 1.25, W = 1.75),
                y = c(X = 1, Y = 1, Z = 2, W = 2))
)

rule2_g_modified <- dagify(
  Y ~ X + W,
  Z ~ X + W,
  coords = list(x = c(X = 1, Y = 2, Z = 1.25, W = 1.75),
                y = c(X = 1, Y = 1, Z = 2, W = 2))
) 
```

```{r}
plot_rule2_g <- ggplot(rule2_g, aes(x = x, y = y, 
                                    xend = xend, yend = yend)) +
  geom_dag_edges() +
  geom_dag_point(color = "grey80", size = 10) +
  geom_dag_text() +
  labs(title = TeX("$G$"),
       subtitle = "Original DAG") +
  theme_do_calc()

plot_rule2_modified <- ggplot(rule2_g_modified, 
                              aes(x = x, y = y, 
                                  xend = xend, yend = yend)) +
  geom_dag_edges() +
  geom_dag_point(color = "grey80", size = 10) +
  geom_dag_text() +
  labs(title = TeX("$G_{\\bar{X}, \\underline{Z}}$"),
       subtitle = "DAG with arrows *into* X and *out of* Z deleted") +
  theme_do_calc()

plot_rule2_g | plot_rule2_modified
```

Okay. Our goal here is to check if we can treat $\mbox{do}(z)$ like a regular observational $z$. We can legally do this if Y and Z are d-separated in that modified graph, after accounting for both W and X, or $Y \bot Z|W, X$. And that is indeed the case! There’s no direct arrow connecting Y and Z in the modified graph, and once we condition on (or account for) W and X, no pathways between Y and Z are active — Y and Z are independent and d-separated. We can confirm this with code:

```{r}
impliedConditionalIndependencies(rule2_g)
```

```{r}
impliedConditionalIndependencies(rule2_g_modified)
```

The second independency there is that $Y \bot Z|W, X$, which is exactly what we want to see. We can thus legally transform $\mbox{do}(z)$ to $z$:

\[
P(y | \mbox{do}(z),\mbox{do}(x),w) = P(y | z,\mbox{do}(x),w)
\]

What’s really neat is that Rule 2 is a generalized version of the backdoor criterion. More on that below after we explore Rule 3.

### Rule 3: Ignoring interventions

Rule 3 is the trickiest of the three, conceptually. It tells us when we can completely remove a $\mbox{do}(\cdot)$ expression rather than converting it to an observed quantity. Here it is in all its mathy glory:

\[
P(y | \mbox{do}(z),\mbox{do}(x),w) = P(y | \mbox{do}(x),w) \quad \mbox{if} (Y \bot Z|W, X) G_{\overline{X},\overline{Z(W)}}
\]

In simpler language, this means that we can ignore an intervention (or a $\mbox{do}(\cdot)$ expression) if it doesn’t influence the outcome through any uncontrolled path — we can remove $\mbox{do}(z)$ if there is no causal association (or no unblocked causal paths) flowing from Z to Y.

This rule is tricky, though, because it depends on where the Z node (i.e. the intervention we want to get rid of) appears in the graph. Note the notation for the modified graph here. With the other rules, we used things like $G_{\overline{X}}$ or $G_{\underline{Z}}$ to remove arrows into and out of specific nodes in the modified graph. Here, though, we have the strange $G_{\overline{X},\overline{Z(W)}}$. This Z(W) is weird! It means "any Z node that isn't an ancestor of W". We thus only delete arrows going into a Z node in the modified graph if that Z node doesn’t precede W.

Here's one version of what that could look like graphically:

```{r}
rule3_g <- dagify(
  Y ~ X + W,
  W ~ Z,
  Z ~ X,
  coords = list(x = c(X = 1, Y = 2, Z = 1.25, W = 1.5),
                y = c(X = 1, Y = 1, Z = 2, W = 1.75))
)
```

```{r}
plot_rule3_g <- ggplot(rule3_g, 
                       aes(x = x, y = y, 
                           xend = xend, yend = yend)) +
  geom_dag_edges() +
  geom_dag_point(color = "grey80", size = 10) +
  geom_dag_text() +
  labs(title = TeX("$G$"),
       subtitle = "Original DAG") +
  theme_do_calc()

plot_rule3_g_modified <- ggplot(rule3_g, 
                                aes(x = x, y = y, 
                                    xend = xend, yend = yend)) +
  geom_dag_edges() +
  geom_dag_point(color = "grey80", size = 10) +
  geom_dag_text() +
  labs(title = TeX("$G_{\\bar{X}, \\bar{Z(W)}}$"),
       subtitle = "DAG with arrows *into* Z deleted as long as Z isn't an<br>ancestor of W + all arrows *into* X deleted") +
  theme_do_calc()

plot_rule3_g | plot_rule3_g_modified
```

Notice how these two graphs are identical. Because we only delete arrows going into Z if Z is not an ancestor of W, in this case $G = G_{\overline{X},\overline{Z(W)}}$. 

Remember that our original goal is to get rid of $\mbox{do}(z)$, which we can legally do if Y and Z are d-separated and independent in our modified graph, or if $Y \bot Z|W, X$. That is once again indeed the case here: there's no direct arrow between Y and Z, and if we condition on W and X, there's no way to pass association between Y and Z, meaning that Y and Z are d-separated. Let's confirm it with code:

```{r}
impliedConditionalIndependencies(rule3_g)
```

That second independency is our $Y \bot Z|W, X$, so we can safely eliminate $\mbox{do}(z)$ from the equation. We can ignore it because it doesn’t influence the outcome Y through any possible path. Goodbye $\mbox{do}(z)$!:

\[
P(y | \mbox{do}(z),\mbox{do}(x),w) = P(y | \mbox{do}(x),w)
\]

In this case, the alternative graph $G_{\overline{X},\overline{Z(W)}}$ was the same as the original graph because of the location of Z — Z was an ancestor of W, so we didn’t delete any arrows. If Z is not an ancestor, though, we get to actually modify the graph. For instance, consider this DAG:

```{r}
rule3_g_alt <- dagify(
  Y ~ X + W,
  Z ~ W,
  X ~ Z,
  coords = list(x = c(X = 1, Y = 2, Z = 1.25, W = 1.5),
                y = c(X = 1, Y = 1, Z = 2, W = 1.75))
)

rule3_g_alt_modified <- dagify(
  Y ~ X + W,
  Z ~ 0,
  X ~ 0,
  coords = list(x = c(X = 1, Y = 2, Z = 1.25, W = 1.5),
                y = c(X = 1, Y = 1, Z = 2, W = 1.75))
) 
```

```{r}
plot_rule3_g_alt <- ggplot(rule3_g_alt, 
                           aes(x = x, y = y, 
                               xend = xend, yend = yend)) +
  geom_dag_edges() +
  geom_dag_point(color = "grey80", size = 10) +
  geom_dag_text() +
  labs(title = TeX("$G$"),
       subtitle = "Original DAG") +
  theme_do_calc()

plot_rule3_g_alt_modified <- ggplot(rule3_g_alt_modified, 
                                    aes(x = x, y = y, 
                                        xend = xend, yend = yend)) +
  geom_dag_edges() +
  geom_dag_point(color = "grey80", size = 10) +
  geom_dag_text() +
  labs(title = TeX("$G_{\\bar{X}, \\bar{Z(W)}}$"),
       subtitle = "DAG with arrows *into* Z deleted as long as Z isn't an<br>ancestor of W + all arrows *into* X deleted") +
  theme_do_calc()

plot_rule3_g_alt | plot_rule3_g_alt_modified
```

Phew. In this case, our DAG surgery for making the modified graph $G_{\overline{X},\overline{Z(W)}}$ actually ended up completely d-separating Z from all nodes. Because Z isn’t an ancestor of W (but is instead a descendant), we get to delete arrows going into it, and we get to delete arrows going into X as well. We can remove $\mbox{do}(z)$ from the equation as long as $Y \bot Z|W, X$ in this modified graph. That is most definitely the case here. And once again, code confirms it (ignore the 0s here - they’re only there so that the DAG plots correctly):

```{r}
impliedConditionalIndependencies(rule3_g_alt_modified)
```

And once again, we can legally get rid of $\mbox{do}(z)$:

\[
P(y | \mbox{do}(z),\mbox{do}(x),w) = P(y | \mbox{do}(x),w)
\]

### Summary

Phew. Let’s look back at the three main rules and add their corresponding mathy versions, which should make more sense now:

- Rule 1: Decide if we can ignore an observation
\[
P(y|z,\mbox{do}(x)),w) = P(y|\mbox{do}(x),w) \quad \mbox{if} (Y \bot Z|W, X) G_{\overline{X}}
\]

- Rule 2: Decide if we can treat an intervention as an observation
\[
P(y | \mbox{do}(z),\mbox{do}(x),w) = P(y | z,\mbox{do}(x),w) \quad \mbox{if} (Y \bot Z|W, X) G_{\overline{X},\underline{Z}}
\]

- Rule 3: Decide if we can ignore an intervention
\[
P(y | \mbox{do}(z),\mbox{do}(x),w) = P(y | \mbox{do}(x),w) \quad \mbox{if} (Y \bot Z|W, X) G_{\overline{X},\overline{Z(W)}}
\]

## Deriving the backdoor adjustment formula from do-calculus rules

That was a lot of math, but hopefully each of these do-calculus rules make sense in isolation now. Now that I finally understand what each of these are doing, we can apply these rules to see where the pre-derived / canned backdoor adjustment formula comes from. Somehow by applying these rules, we can transform the left-hand side of this formula into the do-free right-hand side:

\[
P(y|\mbox{do}(x)) = P(y|x,z) \times P(z)
\]

Let’s go through the derivation of the backdoor adjustment formula step-by-step to see how it works. We’ll use this super simple DAG that shows the causal effect of treatment X on outcome Y, confounded by Z:

```{r}
backdoor_g <- dagify(
  Y ~ X + Z,
  X ~ Z,
  coords = list(x = c(Y = 2, X = 1, Z = 1.5),
                y = c(Y = 1, X = 1, Z = 2))
)

plot_backdoor_g <- ggplot(backdoor_g, aes(x = x, y = y, 
                                          xend = xend, yend = yend)) +
  geom_dag_edges() +
  geom_dag_point(color = "grey80", size = 10) +
  geom_dag_text() +
  labs(title = TeX("$G$"),
       subtitle = "Basic backdoor confounding") +
  theme_do_calc()
plot_backdoor_g
```

### Marginalizing across $z$

We’re interested in the causal effect of X on Y, or $P(y|\mbox{do}(x))$. If this were an experiment like a randomized controlled trial, we’d be able to delete all arrows going into X, which would remove all confounding from Z and allow us to measure the exact causal effect of X on Y. However, with observational data, we can’t delete arrows like that. But, we can condition the X &rarr; Y relationship on Z, given that it influences both X and Y.

We thus need to calculate the joint probability of $P(y|\mbox{do}(x))$ across all values of Z. Using the rules of probability marginalization[^1] and the chain rule[^2] for joint probabilities, we can write this joint probability like so:

\[
P(y|\mbox{do}(x)) = P(y|\color{Red}{\mbox{do}(x)},z) \times P(\color{Purple}{\mbox{do}(x)})
\]

The right-hand side of that equation is what we want to be able to estimate using only observational data, but right now it has two $\mbox{do}(\cdot)$ operators in it, marked in red and purple:

### Applying Rule 2

First let’s get rid of the red $\color{Red}{\mbox{do}(x)}$ that’s in $P(y|\color{Red}{\mbox{do}(x)},z)$. This chunk of the equation involves all three variables: treatment, outcome, and confounder. Accordingly, we don’t really want to ignore any of these variables by using something like Rule 1 or Rule 3. Instead, we can try to treat that $\color{Red}{\mbox{do}(x)}$ as an observational $\color{Red}{x}$ using Rule 2.

According to Rule 2, we can treat an interventional $\mbox{do}(x)$ operator as observational if we meet specific criteria in a modified graph where we remove all arrows out of X:

\[
P(y|\color{Red}{\mbox{do}(x)},z) = P(y | \color{Red}{x},z) \quad \mbox{if}(Y \bot X |Z) G_{\underline{X}}
\]

Here’s the modified $G_{\underline{X}}$ graph: 

```{r}
backdoor_g_underline_x <- dagify(
  Y ~ Z,
  X ~ Z,
  coords = list(x = c(Y = 2, X = 1, Z = 1.5),
                y = c(Y = 1, X = 1, Z = 2))
)

plot_backdoor_g_underline_x <- ggplot(backdoor_g_underline_x, 
                                      aes(x = x, y = y, 
                                          xend = xend, yend = yend)) +
  geom_dag_edges() +
  geom_dag_point(color = "grey80", size = 10) +
  geom_dag_text() +
  labs(title = TeX("$G_{\\underline{X}}$"),
       subtitle = "DAG with arrows *out of* X deleted") +
  theme_do_calc()

plot_backdoor_g | plot_backdoor_g_underline_x
```

Following Rule 2, we can treat $\color{Red}{\mbox{do}(x)}$ like a regular observational $\color{Red}{x}$ as long as X and Y are d-separated in this modified $G_{\underline{X}}$ graph when conditioning on Z. And that is indeed the case: there’s no direct arrow between X and Y, and by conditioning on Z, there’s no active pathway between X and Y through Z. Let’s see if code backs us up:

```{r}
impliedConditionalIndependencies(backdoor_g_underline_x)
```

Perfect! Because $Y \bot X |Z$, we can treat $\color{Red}{\mbox{do}(x)}$ like $\color{Red}{x}$.



[^1]: https://en.wikipedia.org/wiki/Marginal_distribution
[^2]: https://en.wikipedia.org/wiki/Chain_rule_(probability)