---
title: "Forage Or Farm Causal Diagram"
author: "Simon Brewer"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries and whatnot

```{r warning=FALSE, message=FALSE}
library(tidyverse)
# library(broom)
library(patchwork)
library(scales)
library(dagitty)
library(ggdag)
library(latex2exp)  # Easily convert LaTeX into arcane plotmath expressions
library(ggtext)     # Use markdown in ggplot labels

# Create a cleaner serifed theme to use throughout
theme_do_calc <- function() {
  theme_dag(base_family = "Times New Roman") +
    theme(plot.title = element_text(size = rel(1.5)),
          plot.subtitle = element_markdown())
}

# Make all geom_dag_text() layers use these settings automatically
update_geom_defaults(ggdag:::GeomDagText, list(family = "Times New Roman", 
                                               fontface = "bold",
                                               color = "black"))
```

## Model 1

Model linking:

- climate 
- agriculture
- population
- resource availability
- resource ownership

```{r}
fof_dag <- dagify(avail ~ density,
                  own ~ avail,
                  agri ~ climate + avail + own,
                  coords = list(x = c(climate = 1, density = 2, 
                                      avail = 2, own = 3,
                                      agri = 2),
                                y = c(climate = 3, density = 3, 
                                      avail = 2, own = 2,
                                      agri = 1)))

ggdag(fof_dag) +
  theme_dag()
```

```{r}
impliedConditionalIndependencies(fof_dag)
```

## Climate

Adjustment set: shows what adjustment is necessary to obtain the causal link between input and outcome

```{r}
ggdag_adjustment_set(fof_dag,
  exposure = "climate",
  outcome = "agri") +
  theme_dag()
```

## Population density

Adjustment set: shows what adjustment is necessary to obtain the causal link between input and outcome

```{r}
ggdag_adjustment_set(fof_dag,
  exposure = "density",
  outcome = "agri") +
  theme_dag()
```

## Population density

Adjustment set: shows what adjustment is necessary to obtain the causal link between input and outcome

```{r}
ggdag_adjustment_set(fof_dag,
  exposure = "avail",
  outcome = "agri") +
  theme_dag()
```

## Ownership

Adjustment set: shows what adjustment is necessary to obtain the causal link between input and outcome

```{r}
ggdag_adjustment_set(fof_dag,
  exposure = "own",
  outcome = "agri") +
  theme_dag()
```

# Model 2

Model linking:

- climate 
- agriculture
- population
- resource availability
- resource ownership

```{r}
fof_dag <- dagify(avail ~ density + climate,
                  own ~ avail,
                  agri ~ climate + avail + own,
                  coords = list(x = c(climate = 1, density = 2, 
                                      avail = 2, own = 3,
                                      agri = 2),
                                y = c(climate = 3, density = 3, 
                                      avail = 2, own = 2,
                                      agri = 1)))

ggdag(fof_dag) +
  theme_dag()
```

```{r}
impliedConditionalIndependencies(fof_dag)
```

## Climate

Adjustment set: shows what adjustment is necessary to obtain the causal link between input and outcome

```{r}
ggdag_adjustment_set(fof_dag,
  exposure = "climate",
  outcome = "agri") +
  theme_dag()
```

## Population density

Adjustment set: shows what adjustment is necessary to obtain the causal link between input and outcome

```{r}
ggdag_adjustment_set(fof_dag,
  exposure = "density",
  outcome = "agri") +
  theme_dag()
```

## Availability

Adjustment set: shows what adjustment is necessary to obtain the causal link between input and outcome

```{r}
ggdag_adjustment_set(fof_dag,
  exposure = "avail",
  outcome = "agri") +
  theme_dag()
```

## Ownership

Adjustment set: shows what adjustment is necessary to obtain the causal link between input and outcome

```{r}
ggdag_adjustment_set(fof_dag,
  exposure = "own",
  outcome = "agri") +
  theme_dag()
```

