---
title: "Causal Frontdoor Examples"
author: "Simon Brewer"
date: "5/21/2023"
output: html_document
---

```{r setup, include=FALSE}
set.seed(123)
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries and whatnot

```{r warning=FALSE, message=FALSE}
library(tidyverse)
# library(broom)
library(patchwork)
library(scales)
library(dagitty)
library(ggdag)
library(latex2exp)  # Easily convert LaTeX into arcane plotmath expressions
library(ggtext)     # Use markdown in ggplot labels

# Create a cleaner serifed theme to use throughout
theme_do_calc <- function() {
  theme_dag(base_family = "Times New Roman") +
    theme(plot.title = element_text(size = rel(1.5)),
          plot.subtitle = element_markdown())
}

# Make all geom_dag_text() layers use these settings automatically
update_geom_defaults(ggdag:::GeomDagText, list(family = "Times New Roman", 
                                               fontface = "bold",
                                               color = "black"))
```

Examples taken from: https://arelbundock.com/posts/frontdoor/

## Front door path

The front door path is used to replace a back door path when there is an unknown (suspected) confounder ($u$). But this has the condition that there needs to be at least one mediating factors between the intervention and the outcome. So this won't work:

```{r}
fd_dag <- dagify(x ~ u,
                 y ~ x + u,
                 coords = list(x = c(x = 1, y = 3, u = 2),
                               y = c(x = 1, y = 1, u = 2)),
                 exposure = "x",
                 outcome = "y")

ggdag(fd_dag) +
  theme_dag()
```

But this will

```{r}
fd_dag <- dagify(x ~ u,
                 z ~ x,
                 y ~ z + u,
                 coords = list(x = c(x = 1, y = 3, u = 2, z = 2),
                               y = c(x = 1, y = 1, u = 2, z = 1)),
                 exposure = "x",
                 outcome = "y")

ggdag(fd_dag) +
  theme_dag()
```

In this DAG:

- `x` is the intervention or covariate of causal interest
- `y` is the outcome
- `u` is an unknown confounder
- `z` is a mediating variable (mediates `y ~ x`)

Our goal is to estimate $P(Y|\mbox{do}(X))$. Unfortunately, this relationship between $X$ and $Y$ is confounded via a backdoor path with an unobserved variable $U$. The path is $X \leftarrow U \rightarrow Y$. Given the backdoor path, we can't estimate the causal relationship (or we can but it will be biased). 

We can instead, estimate the effect of $X$ on $Y$ indirectly through the front door path ($X \rightarrow Z \rightarrow Y$), by decomposing it into two links:

- The effect of $X$ and $Z$
- The effect of $Z$ on $Y$ given $X$

There are three steps here

- Estimate $P(Z|\mbox{do}(X))$
- Estimate $P(Y|\mbox{do}(Z), X)$
- Combine the two estimates

## Simulated data

All binary:

- `U`: $P=0.2$
- `X`: $P=0.1 + U \times 0.6$
- `Z`: $P=0.3 + X \times 0.5$
- `Y`: $P=0.1 + U \times 0.3 + Z \times 0.5$

```{r}
N = 1e5
U = rbinom(N, 1, prob = .2)
X = rbinom(N, 1, prob = .1 + U * .6)
Z = rbinom(N, 1, prob = .3 + X * .5)
Y = rbinom(N, 1, prob = .1 + U * .3 + Z * .5)
dat = data.frame(X, Z, Y)
```

The true effect of `X` on `Y` is $0.5 \times 0.5 = 0.25$. In this simple case, we should be able to obtain this effect by creating the two models (one for `Z`, one for `Y`) and multiplying the coefficients. First, start with a naive model (i.e. no accounting for any paths or mediators):

```{r}
fit <- lm(Y ~ X, dat)
summary(fit)
```

Which results in a biased estimate of `r round(coef(fit)["X"], 3)`. We can also try a model that assumes an independent effect of `X` and `Y`:

```{r}
fit <- lm(Y ~ X + Z, dat)
summary(fit)
```

And the effect is still biased (`r round(coef(fit)["X"], 3)`). 

### Front-door adjustment

First we estimate the effect of `X` on `Z`. Since there is no open backdoor, we do not need to control for other variables:

```{r}
step1 = lm(Z ~ X, dat)
```

Then, we estimate the effect of `Z` on `Y` while controlling for `X` to close the backdoor:

```{r}
step2 = lm(Y ~ Z + X, dat)
```

Finally, we combine the two estimates by multiplication:

```{r}
coef(step1)["X"] * coef(step2)["Z"] 
```

## Frontdoor Algebra 

We require the following assumptions (https://stats.stackexchange.com/questions/312992/causal-effect-by-back-door-and-front-door-adjustments):

- Full mediation: there is no directed path from `X` to `Y`, except through `Z`
- Unconfoundedness 1: There is no open backdoor from `X` to `Z`
- Unconfoundedness 2: All backdoors from `Z` to `Y` are blocked by `X`.

Step 1: Under assumption 2, the relationship between `X` and `Z` is not confounded. See DAG at the top - there's no way to get from `X` to `Z` except the direct path. The other path is blocked by a collider at `Y`. As a result, we have:
\[
P(Z|\mbox{do}(x)) = P(Z|X)
\]

Step 2: In contrast, the relationship between `Z` and `Y` is confounded ($Z \leftarrow X \leftarrow U \rightarrow Y$. If we adjust for `X`, this blocks this path, and allows an unbiased estimate of the effect of `Z`. The backdoor formula is:

\[
P(Y|\mbox{do}(Z)) = \sum_X P(Y|X,Z)P(X)
\]

$$
\begin{aligned}
  P(Y|\mbox{do}(Z)) & = \sum_Z P(Y|\mbox{do}(X),Z)P(Z|\mbox{do}(X)) \\
                    & = \sum_Z P(Y|\mbox{do}(Z))P(Z|\mbox{do}(X)) \\
                    & = \sum_Z P(Y|\mbox{do}(Z))P(Z|X) \\
                    & = \sum_Z \sum_X P(Y|X,Z)P(X)P(Z|X) \\
                    & = \sum_Z P(Z|X) \sum_X P(Y|X,Z)P(X) 
\end{aligned}
$$

Equation (1) conditions on `Z` and sums over its values. Equation (2) is allowed because the effect of `X` is entirely mediated by `Z`, and because `X` blocks the backdoor path from `Z` to `Y`. Intuitively, if we manipulate `Z`, it no longer matters what happened to `X`. Equation (3) is allowed because the relationship between `X` and `Z` is unconfounded. Equation (4) applies the backdoor adjustment formula to estimate the effect of `Z` on `Y` by conditioning on `X`. 

Equation (5) is the frontdoor adjustment formula. The left part is the effect of `X` on `Z`. The right part is the effect of `Z` on `Y`.

## **dosearch**

The **dosearch** package for R includes an algorithm that can automatically apply the rules of do-calculus to convert DAGs to adjustment formulas. Instead of manipulating equations ourselves like we did above, we can simply call the `dosearch` function to obtain the frontdoor adjustment formula automatically:

```{r}
library('dosearch')

data1 = "P(X, Y, Z)"

query1 = "P(Y | do(X))"

graph1 = "
U -> X
U -> Y
X -> Z
Z -> Y
"

# compute
frontdoor = dosearch(data1, query1, graph1)

# convert to Rmarkdown equation
cat(paste("$$", frontdoor$formula, "$$"))
```

$$`r frontdoor$formula`$$

