---
title: "Causal Mediation Examples"
author: "Simon Brewer"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
set.seed(42)
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries and whatnot

```{r warning=FALSE, message=FALSE}
library(tidyverse)
# library(broom)
library(patchwork)
library(scales)
library(dagitty)
library(ggdag)
library(latex2exp)  # Easily convert LaTeX into arcane plotmath expressions
library(ggtext)     # Use markdown in ggplot labels
library(mediation)
library(lavaan)
library(lavaanPlot)

# Create a cleaner serifed theme to use throughout
theme_do_calc <- function() {
  theme_dag(base_family = "Times New Roman") +
    theme(plot.title = element_text(size = rel(1.5)),
          plot.subtitle = element_markdown())
}

# Make all geom_dag_text() layers use these settings automatically
update_geom_defaults(ggdag:::GeomDagText, list(family = "Times New Roman", 
                                               fontface = "bold",
                                               color = "black"))
```

## Examples

In all the examples that follow, we are trying to estimate a causal relationship between `x` and `y`. 

## Example 1: Single mediator

This represents an indirect path from intervention to outcome via a mediator (`x` &rarr; `z` &rarr; `y`), as well as a direct path (`x` &rarr; `y`)

```{r}
med_dag <- dagify(z ~ x,
                  y ~ x + z,
                  coords = list(x = c(x = 1, y = 3, z = 2),
                                y = c(x = 1, y = 1, z = 2)),
                  exposure = "x",
                  outcome = "y")

ggdag(med_dag) +
  theme_dag()
```

### Simulated data

- $x \sim N(5, 1)$
- $z = \beta_{xz} \times x + e$
- $y = \beta_{xy} \times x + \beta_{zy} \times z + e$

```{r}
n <- 1000

b_xz = 4
b_xy = 2
b_zy = -1.5

x <- rnorm(n, 5, 1)
z <- x * b_xz + rnorm(n, 0, 2)
y <- x * b_xy + z * b_zy + rnorm(n, 0, 2)
df <- data.frame(x, y, z)
```

```{r}
pairs(df)
```

### Mediation

#### Step 1: Total effect

```{r}
fit1 <- lm(y ~ x, df)
summary(fit1)
```

The total effect is given by the product of the coefficients of the indirect path plus the coefficient of the direct path: $TE = \beta_{xz} \times \beta_{zy} + \beta_{xy}$

The expected value for this is `r b_xz` times `r b_zy` plus `r b_xy` = `r b_xz * b_zy + b_xy`, and we obtain a value of `r round(coef(fit1)['x'], 2)`. 

#### Step 2: Effect of IV on mediator

```{r}
fit2 <- lm(z ~ x, df)
summary(fit2)
```

With no confounders, this is simply $IVE = \beta_{xz} \times x$.
The expected value for this is `r b_xz`, and we obtain a value of `r round(coef(fit2)['x'], 2)`. 

#### Step 3: Effect of mediator on DV

To get the effect of `z` xon `y` we need to account for `x` as it is a fork or common cause

```{r}
ggdag_adjustment_set(med_dag, 
                     exposure = 'z',
                     outcome = 'y') + 
  theme_dag()
```

```{r}
fit3 <- lm(y ~ x + z, df)
summary(fit3)
```

The mediator (`z` &rarr; `y`) is `r round(coef(fit3)['z'], 2)`, which should be close to the known effect of `r b_zy` (at least within CIs). As the direct effect in this model (`x` &rarr; `y`) is significant, this is *partial mediation*.

#### Step 4: Estimate mediation effects

```{r}
res = mediation::mediate(model.m = fit2, 
                         model.y = fit3, 
                         treat = 'x',
                         mediator = 'z',
                         boot = TRUE)
summary(res)
```

- `ACME` (average causal mediation effect). The indirect effect of the independent variable / treatment (`x`) on the outcome (`y`), in other words, the effect that passes through the mediator `z`. The value estimated here is `r round(res$d0, 3)`. This is the effect of the treatment on the mediator (`r round(coef(fit2)['x'], 2)`) from step 2 times the mediator's effect on the outcome from step 3 (`r round(coef(fit3)['z'], 2)`)
- `ADE` stands for average direct effects. This is the direct effect of the treatment on the outcome, and is the same as the `x` coefficient from step 3
- `Total Effect` stands for the total effect (direct + indirect). This is the same  as the coefficient from model 1 (`r round(coef(fit1)['x'], 2)`)
- We can also get it by simply adding the ACME (`r round(res$d0, 3)`) and the ADE (`r round(res$z0, 3)`) to get the total effect of `r round(res$tau.coef, 3)` 
- `Prop. Mediated` describes the proportion of the effect of the IV on the DV that goes through the mediator. This is calculated by dividing the ACME (`r round(res$d0, 3)`) by the total effect (`r round(res$tau.coef, 3)`) to get a value of `r round(res$n0, 3)`. 

## Example 2: Two mediators

This represents an indirect path from intervention to outcome via two independent mediators (`x` &rarr; `z1` &rarr; `y`) and (`x` &rarr; `z2` &rarr; `y`)

```{r}
med_dag <- dagify(z1 ~ x,
                  z2 ~ x,
                  y ~ x + z1 + z2,
                  coords = list(x = c(x = 1, y = 3, z1 = 2, z2 = 2),
                                y = c(x = 1, y = 1, z1 = 2, z2 = 0)),
                  exposure = "x",
                  outcome = "y")

ggdag(med_dag) +
  theme_dag()
```

### Simulated data

- $x \sim N(5, 1)$
- $z_1 = \beta_{xz_1} \times x + e$
- $z_2 = \beta_{xz_2} \times x + e$
- $y = \beta_{xy} \times x + \beta_{z_1y} \times z_1 + \beta_{z_2y} \times z_2 + e$

```{r}
n <- 10000

b_xz1 = 4
b_xz2 = -3
b_xy = 2
b_z1y = -1.5
b_z2y = -0.5

x <- rnorm(n, 5, 1)
z1 <- x * b_xz1 + rnorm(n, 0, 2)
z2 <- x * b_xz2 + rnorm(n, 0, 2)
y <- x * b_xy + z1 * b_z1y + z2 * b_z2y + rnorm(n, 0, 2)
df <- data.frame(x, y, z1, z2)
```

```{r}
pairs(df)
```

### Mediation

#### Step 1: Total effect

```{r}
fit1 <- lm(y ~ x, df)
summary(fit1)
```

The total effect is given by the product of the coefficients of the indirect path plus the coefficient of the direct path: $TE = \beta_{xz_1} \times \beta_{z_1y} + \beta_{xz_2} \times \beta_{z_2y} + \beta_{xy}$

The expected value for this is `r b_xz1` $\times$ `r b_z1y` $+$ `r b_xz2` $\times$ `r b_z2y` $+$ `r b_xy` = `r b_xz1 * b_z1y + b_xz2 * b_z2y + b_xy`, and we obtain a value of `r round(coef(fit1)['x'], 2)`. 

#### Step 2: Effect of IV on mediator

- `z1`

```{r}
fit2a <- lm(z1 ~ x, df)
summary(fit2a)
```

The expected value for this is `r b_xz1`, and we obtain a value of `r round(coef(fit2a)['x'], 2)`. 

- `z2`

```{r}
fit2b <- lm(z2 ~ x, df)
summary(fit2b)
```

The expected value for this is `r b_xz2`, and we obtain a value of `r round(coef(fit2b)['x'], 2)`. 

#### Step 3: Effect of mediators on DV

To get the effect of `z1` and `z2` on `y` we need to account for `x` as it is a fork or common cause (not shown here)

```{r}
fit3 <- lm(y ~ x + z1 + z2, df)
summary(fit3)
```

The mediator (`z1` &rarr; `y`) is `r round(coef(fit3)['z1'], 2)`, which should be close to the known effect of `r b_z1y` (at least within CIs). 

The mediator (`z2` &rarr; `y`) is `r round(coef(fit3)['z2'], 2)`, which should be close to the known effect of `r b_z2y` (at least within CIs). 

The direct effect in this model (`x` &rarr; `y`) is `r round(coef(fit3)['x'], 2)`. As this is significant, this is *partial mediation*.

#### Step 4: Estimate mediation effects

The **mediation** package won't work with two mediators (**psych** does, see below for code)


##### ACME 

`ACME` (average causal mediation effect). The indirect effect of the independent variable / treatment (`x`) on the outcome (`y`), in other words, the effect that passes through the two mediators `z1` and `z1`

This is the sum of the two individual mediator effects:

\[
ACME = (\beta_{xz_1} \times \beta_{z_1y}) + (\beta_{xz_2} \times \beta_{z_2y})
\]

```{r}
ACME = (coef(fit2a)['x'] * coef(fit3)['z1']) + (coef(fit2b)['x'] * coef(fit3)['z2'])
ACME
```

Expected ACME is
```{r}
ACMEe = (b_xz1 * b_z1y) + (b_xz2 * b_z2y)
ACMEe
```

##### ADE

`ADE` stands for average direct effects. This is the direct effect of the treatment on the outcome, and is `x` coefficient from step 3

```{r}
ADE = coef(fit3)['x']
ADE
```

##### Total effect

`Total Effect` stands for the total effect (direct + indirect). This is the same  as the coefficient from model 1 (`r round(coef(fit1)['x'], 2)`). We can also get it by simply adding the ACME and the ADE

```{r}
ACME + ADE
```


```{r}
load("example2.RData")
psych::mediate.diagram(res)
#psych::mediate(y ~ x + (z1) + (z2), data = df)
```

## Example 3: Two stage mediation

This represents an indirect path from intervention to outcome via two dependent mediators (`x` &rarr; `z1` &rarr; `z2` &rarr; `y`)

```{r}
med_dag <- dagify(z1 ~ x,
                  z2 ~ z1,
                  y ~ x + z2,
                  coords = list(x = c(x = 1, y = 4, z1 = 2, z2 = 3),
                                y = c(x = 1, y = 1, z1 = 2, z2 = 2)),
                  exposure = "x",
                  outcome = "y")

ggdag(med_dag) +
  theme_dag()
```

### Simulated data

- $x \sim N(5, 1)$
- $z_1 = \beta_{xz_1} \times x + e$
- $z_2 = \beta_{z_1z_2} \times z_1 + e$
- $y = \beta_{xy} \times x + \beta_{z_2y} \times z_2 + e$

```{r}
n <- 1000

b_xz1 = 10
b_z1z2 = -5
b_xy = 2.5
b_z2y = -0.5

x <- rnorm(n, 5, 1)
z1 <- x * b_xz1 + rnorm(n, 0, 5)
z2 <- z1 * b_z1z2 + rnorm(n, 0, 5)
y <- x * b_xy + z2 * b_z2y + rnorm(n, 0, 5)
df <- data.frame(x, y, z1, z2)
```

```{r}
pairs(df)
```

### Mediation

#### Step 1: Total effect

```{r}
fit1 <- lm(y ~ x, df)
summary(fit1)
```

#### Step 2: Effect of IV on mediator (`z1`)

- `z1`

```{r}
fit2 <- lm(z1 ~ x, df)
summary(fit2)
```

The expected value for this is `r b_xz1`, and we obtain a value of `r round(coef(fit2)['x'], 2)`. 

#### Step 3: Effect of mediator (`z1`) on mediator (`z2`)

```{r}
fit3 <- lm(z2 ~ z1, df)
summary(fit3)
```

The expected value for this is `r b_z1z2`, and we obtain a value of `r round(coef(fit3)['z1'], 2)`. 

#### Step 4: Effect of mediator on DV

To get the effect of  `z2` on `y` we need to account for `x` as it is a fork or common cause (not shown here)

```{r}
fit4 <- lm(y ~ x + z2, df)
summary(fit4)
```

The mediator (`z2` &rarr; `y`) is `r round(coef(fit4)['z2'], 2)`, which should be close to the known effect of `r b_z2y` (at least within CIs). 

The direct effect in this model (`x` &rarr; `y`) is `r round(coef(fit4)['x'], 2)`. As this is significant, this is *partial mediation*.

#### Step 5: Estimate mediation effects


##### ACME 

`ACME` (average causal mediation effect). The indirect effect of the independent variable / treatment (`x`) on the outcome (`y`), in other words, the effect that passes through the two mediators `z1` and `z1`

This (I think) is the product of the three individual mediator effects (`x` &rarr; `z1` &rarr; `z2` &rarr; `y`)

\[
ACME = \beta_{xz_1} \times \beta_{z_1z_2} \times \beta_{z_2y}
\]

```{r}
ACME = coef(fit2)['x'] * coef(fit3)['z1'] * coef(fit4)['z2']
ACME
```

Expected ACME is
```{r}
ACMEe = b_xz1 * b_z1z2 * b_z2y
ACMEe
```

##### ADE

`ADE` stands for average direct effects. This is the direct effect of the treatment on the outcome, and is `x` coefficient from step 3

```{r}
ADE = coef(fit4)['x']
ADE
```

##### Total effect

`Total Effect` stands for the total effect (direct + indirect). This is the same  as the coefficient from model 1 (`r round(coef(fit1)['x'], 2)`). We can also get it by simply adding the ACME and the ADE

```{r}
ACME + ADE
```

#### **lavaan**

```{r}
model <- "
  z1 ~ b_xz1 * x
  z2 ~ b_z1z2 * z1
  y ~  b_xy * x + b_z2y * z2
  ind_eff := b_xz1 * b_z1z2 * b_z2y
  total := b_xy + (b_xz1 * b_z1z2 * b_z2y)
"
```

```{r}
# fit <- lavaan::sem(model = model, data = df, se = "boot", bootstrap = 5000)
fit <- lavaan::sem(model = model, data = df)
lavaanPlot(model = fit)
```

```{r}
summary(fit)
```
