---
title: "Causal Mediation Examples"
author: "Simon Brewer"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
set.seed(42)
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries and whatnot

```{r warning=FALSE, message=FALSE}
library(tidyverse)
# library(broom)
library(patchwork)
library(scales)
library(dagitty)
library(ggdag)
library(latex2exp)  # Easily convert LaTeX into arcane plotmath expressions
library(ggtext)     # Use markdown in ggplot labels
library(mediation)
library(lavaan)
library(lavaanPlot)
```

## Examples

In all the examples that follow, we are trying to estimate a causal relationship between `x` and `y`. 

## Example 1: Single mediator

This represents an indirect path from intervention to outcome via a mediator (`x` &rarr; `z` &rarr; `y`), as well as a direct path (`x` &rarr; `y`)

### Simulated data

- $x \sim N(5, 1)$
- $z = \beta_{xz} \times x + e$
- $y = \beta_{xy} \times x + \beta_{zy} \times z + e$

```{r}
n <- 1000

b_xz = 4
b_xy = 2
b_zy = -1.5

x <- rnorm(n, 5, 1)
z <- x * b_xz + rnorm(n, 0, 2)
y <- x * b_xy + z * b_zy + rnorm(n, 0, 2)
df <- data.frame(x, y, z)
```

```{r}
pairs(df)
```

### DAG

```{r}
model <- "
  z ~ b_xz * x
  y ~ b_xy * x + b_zy * z
  ind_eff := b_xz * b_zy
  total := b_xy + (b_xz * b_zy)
"
semfit <- lavaan::sem(model = model, data = df)
lavaanPlot(model = semfit)
```

### Mediation

#### Step 1: Total effect

```{r}
fit1 <- lm(y ~ x, df)
summary(fit1)
```

The total effect is given by the product of the coefficients of the indirect path plus the coefficient of the direct path: $TE = \beta_{xz} \times \beta_{zy} + \beta_{xy}$

The expected value for this is `r b_xz` times `r b_zy` plus `r b_xy` = `r b_xz * b_zy + b_xy`, and we obtain a value of `r round(coef(fit1)['x'], 2)`. 

#### Step 2: Effect of IV on mediator

```{r}
fit2 <- lm(z ~ x, df)
summary(fit2)
```

With no confounders, this is simply $IVE = \beta_{xz} \times x$.
The expected value for this is `r b_xz`, and we obtain a value of `r round(coef(fit2)['x'], 2)`. 

#### Step 3: Effect of mediator on DV

To get the effect of `z` xon `y` we need to account for `x` as it is a fork or common cause

```{r}
fit3 <- lm(y ~ x + z, df)
summary(fit3)
```

The mediator (`z` &rarr; `y`) is `r round(coef(fit3)['z'], 2)`, which should be close to the known effect of `r b_zy` (at least within CIs). As the direct effect in this model (`x` &rarr; `y`) is significant, this is *partial mediation*.

#### Step 4: Estimate mediation effects

```{r}
summary(semfit)
```

##### ACME

`ACME` (average causal mediation effect). The indirect effect of the independent variable / treatment (`x`) on the outcome (`y`), in other words, the effect that passes through the mediator `z`. The value estimated here is `r round(semfit@Fit@est[7], 3)`. This is the effect of the treatment on the mediator (`r round(coef(fit2)['x'], 2)`) from step 2 times the mediator's effect on the outcome from step 3 (`r round(coef(fit3)['z'], 2)`)

```{r}
coef(fit2)['x'] * coef(fit3)['z']
```
