---
title: "Causal Mediation Examples"
author: "Simon Brewer"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
set.seed(42)
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries and whatnot

```{r warning=FALSE, message=FALSE}
library(lavaan)
library(lavaanPlot)
```

## Examples

In all the examples that follow, we are trying to estimate a causal relationship between `x` and `y`. 

## Example 1: Single mediator

This represents an indirect path from intervention to outcome via a mediator (`x` &rarr; `z` &rarr; `y`), as well as a direct path (`x` &rarr; `y`)

### Simulated data

- $x \sim N(5, 1)$
- $z = \beta_{xz} \times x + e$
- $y = \beta_{xy} \times x + \beta_{zy} \times z + e$

```{r}
n <- 1000

b_xz = 4
b_xy = 2
b_zy = -1.5

x <- rnorm(n, 5, 1)
z <- x * b_xz + rnorm(n, 0, 2)
y <- x * b_xy + z * b_zy + rnorm(n, 0, 2)
df <- data.frame(x, y, z)
```

```{r}
pairs(df)
```

### DAG

```{r}
model <- "
  z ~ b_xz * x
  y ~ b_xy * x + b_zy * z
  ind_eff := b_xz * b_zy
  total := b_xy + (b_xz * b_zy)
"
semfit <- lavaan::sem(model = model, data = df)
lavaanPlot(model = semfit)
```

### Mediation

#### Step 1: Total effect

```{r}
fit1 <- lm(y ~ x, df)
summary(fit1)
```

The total effect is given by the product of the coefficients of the indirect path plus the coefficient of the direct path: $TE = \beta_{xz} \times \beta_{zy} + \beta_{xy}$

The expected value for this is `r b_xz` times `r b_zy` plus `r b_xy` = `r b_xz * b_zy + b_xy`, and we obtain a value of `r round(coef(fit1)['x'], 2)`. 

#### Step 2: Effect of IV on mediator

```{r}
fit2 <- lm(z ~ x, df)
summary(fit2)
```

With no confounders, this is simply $IVE = \beta_{xz} \times x$.
The expected value for this is `r b_xz`, and we obtain a value of `r round(coef(fit2)['x'], 2)`. 

#### Step 3: Effect of mediator on DV

To get the effect of `z` xon `y` we need to account for `x` as it is a fork or common cause

```{r}
fit3 <- lm(y ~ x + z, df)
summary(fit3)
```

The mediator (`z` &rarr; `y`) is `r round(coef(fit3)['z'], 2)`, which should be close to the known effect of `r b_zy` (at least within CIs). As the direct effect in this model (`x` &rarr; `y`) is significant, this is *partial mediation*.

#### Step 4: Estimate mediation effects

```{r}
summary(semfit)
```

##### ACME

`ACME` (average causal mediation effect). The indirect effect of the independent variable / treatment (`x`) on the outcome (`y`), in other words, the effect that passes through the mediator `z`. The value estimated here is `r round(semfit@Fit@est[7], 3)`. This is the effect of the treatment on the mediator (`r round(coef(fit2)['x'], 2)`) from step 2 times the mediator's effect on the outcome from step 3 (`r round(coef(fit3)['z'], 2)`)

```{r}
ACME = coef(fit2)['x'] * coef(fit3)['z']
ACME
```

Expected value: `r b_xz * b_zy`

##### ADE

`ADE` stands for average direct effects. This is the direct effect of the treatment on the outcome, and is the same as the `x` coefficient from step 3

```{r}
ADE = coef(fit3)['x']
ADE
```

Expected value: `r b_xy`

##### Total Effect

`Total Effect` stands for the total effect (direct + indirect). This is the same  as the coefficient from model 1 (`r round(coef(fit1)['x'], 2)`). 

Note we can also get this as the sum of ACME and ADE

```{r}
TE = ACME + ADE
TE
```

##### Proportion Mediated 

`Prop. Mediated` describes the proportion of the effect of the IV on the DV that goes through the mediator. This is calculated by dividing the ACME by the total effect

```{r}
ACME / TE
```

## Example 2: Two mediators

This represents an indirect path from intervention to outcome via two independent mediators (`x` &rarr; `z1` &rarr; `y`) and (`x` &rarr; `z2` &rarr; `y`)

### Simulated data

- $x \sim N(5, 1)$
- $z_1 = \beta_{xz_1} \times x + e$
- $z_2 = \beta_{xz_2} \times x + e$
- $y = \beta_{xy} \times x + \beta_{z_1y} \times z_1 + \beta_{z_2y} \times z_2 + e$

```{r}
n <- 10000

b_xz1 = 4
b_xz2 = -3
b_xy = 2
b_z1y = -1.5
b_z2y = -0.5

x <- rnorm(n, 5, 1)
z1 <- x * b_xz1 + rnorm(n, 0, 2)
z2 <- x * b_xz2 + rnorm(n, 0, 2)
y <- x * b_xy + z1 * b_z1y + z2 * b_z2y + rnorm(n, 0, 2)
df <- data.frame(x, y, z1, z2)
```

```{r}
pairs(df)
```

### DAG

```{r}
model <- "
  z1 ~ b_xz1 * x
  z2 ~ b_xz2 * x
  y ~ b_xy * x + b_z1y * z1 + b_z2y * z2
  ind_eff := b_xz1 * b_z1y + b_xz2 * b_z2y
  total := b_xy + (b_xz1 * b_z1y + b_xz2 * b_z2y)
"
semfit <- lavaan::sem(model = model, data = df)
lavaanPlot(model = semfit)
```

### Mediation

#### Step 1: Total effect

```{r}
fit1 <- lm(y ~ x, df)
summary(fit1)
```

The total effect is given by the product of the coefficients of the indirect path plus the coefficient of the direct path: $TE = \beta_{xz_1} \times \beta_{z_1y} + \beta_{xz_2} \times \beta_{z_2y} + \beta_{xy}$

The expected value for this is `r b_xz1` $\times$ `r b_z1y` $+$ `r b_xz2` $\times$ `r b_z2y` $+$ `r b_xy` = `r b_xz1 * b_z1y + b_xz2 * b_z2y + b_xy`, and we obtain a value of `r round(coef(fit1)['x'], 2)`. 

#### Step 2: Effect of IV on mediator

- `z1`

```{r}
fit2a <- lm(z1 ~ x, df)
summary(fit2a)
```

The expected value for this is `r b_xz1`, and we obtain a value of `r round(coef(fit2a)['x'], 2)`. 

- `z2`

```{r}
fit2b <- lm(z2 ~ x, df)
summary(fit2b)
```

The expected value for this is `r b_xz2`, and we obtain a value of `r round(coef(fit2b)['x'], 2)`. 

#### Step 3: Effect of mediators on DV

To get the effect of `z1` and `z2` on `y` we need to account for `x` as it is a fork or common cause (not shown here)

```{r}
fit3 <- lm(y ~ x + z1 + z2, df)
summary(fit3)
```

The mediator (`z1` &rarr; `y`) is `r round(coef(fit3)['z1'], 2)`, which should be close to the known effect of `r b_z1y` (at least within CIs). 

The mediator (`z2` &rarr; `y`) is `r round(coef(fit3)['z2'], 2)`, which should be close to the known effect of `r b_z2y` (at least within CIs). 

The direct effect in this model (`x` &rarr; `y`) is `r round(coef(fit3)['x'], 2)`. The expected value is `r b_xy`. As this is significant, this is *partial mediation*.

#### Step 4: Estimate mediation effects

```{r}
summary(semfit)
```

##### ACME 

`ACME` (average causal mediation effect). The indirect effect of the independent variable / treatment (`x`) on the outcome (`y`), in other words, the effect that passes through the two mediators `z1` and `z1`

This is the sum of the two individual mediator effects:

\[
ACME = (\beta_{xz_1} \times \beta_{z_1y}) + (\beta_{xz_2} \times \beta_{z_2y})
\]

```{r}
ACME = (coef(fit2a)['x'] * coef(fit3)['z1']) + (coef(fit2b)['x'] * coef(fit3)['z2'])
ACME
```

Expected ACME is
```{r}
ACMEe = (b_xz1 * b_z1y) + (b_xz2 * b_z2y)
ACMEe
```

##### ADE

`ADE` stands for average direct effects. This is the direct effect of the treatment on the outcome, and is `x` coefficient from step 3

```{r}
ADE = coef(fit3)['x']
ADE
```

Expected ADE is `r b_xy`

##### Total effect

`Total Effect` stands for the total effect (direct + indirect). This is the same  as the coefficient from model 1 (`r round(coef(fit1)['x'], 2)`). We can also get it by simply adding the ACME and the ADE

```{r}
ACME + ADE
```

## Example 3: Two stage mediation

This represents an indirect path from intervention to outcome via two dependent mediators (`x` &rarr; `z1` &rarr; `z2` &rarr; `y`)

### Simulated data

- $x \sim N(5, 1)$
- $z_1 = \beta_{xz_1} \times x + e$
- $z_2 = \beta_{z_1z_2} \times z_1 + e$
- $y = \beta_{xy} \times x + \beta_{z_2y} \times z_2 + e$

```{r}
n <- 1000

b_xz1 = 10
b_z1z2 = -5
b_xy = 2.5
b_z2y = -0.5

x <- rnorm(n, 5, 1)
z1 <- x * b_xz1 + rnorm(n, 0, 5)
z2 <- z1 * b_z1z2 + rnorm(n, 0, 5)
y <- x * b_xy + z2 * b_z2y + rnorm(n, 0, 5)
df <- data.frame(x, y, z1, z2)
```

```{r}
pairs(df)
```

### DAG

```{r}
model <- "
  z1 ~ b_xz1 * x
  z2 ~ b_z1z2 * z1
  y ~ b_xy * x + b_z2y * z2
  ind_eff := b_xz1 * b_z1z2 * b_z2y
  total := b_xy + (b_xz1 * b_z1z2 * b_z2y)
"
semfit <- lavaan::sem(model = model, data = df)
lavaanPlot(model = semfit)
```

### Mediation

#### Step 1: Total effect

```{r}
fit1 <- lm(y ~ x, df)
summary(fit1)
```

#### Step 2: Effect of IV on mediator (`z1`)

- `z1`

```{r}
fit2 <- lm(z1 ~ x, df)
summary(fit2)
```

The expected value for this is `r b_xz1`, and we obtain a value of `r round(coef(fit2)['x'], 2)`. 

#### Step 3: Effect of mediator (`z1`) on mediator (`z2`)

```{r}
fit3 <- lm(z2 ~ z1, df)
summary(fit3)
```

The expected value for this is `r b_z1z2`, and we obtain a value of `r round(coef(fit3)['z1'], 2)`. 

#### Step 4: Effect of mediator on DV

To get the effect of  `z2` on `y` we need to account for `x` as it is a fork or common cause (not shown here)

```{r}
fit4 <- lm(y ~ x + z2, df)
summary(fit4)
```

The mediator (`z2` &rarr; `y`) is `r round(coef(fit4)['z2'], 2)`, which should be close to the known effect of `r b_z2y` (at least within CIs). 

The direct effect in this model (`x` &rarr; `y`) is `r round(coef(fit4)['x'], 2)`. As this is significant, this is *partial mediation*.

#### Step 5: Estimate mediation effects

```{r}
summary(semfit)
```

##### ACME 

`ACME` (average causal mediation effect). The indirect effect of the independent variable / treatment (`x`) on the outcome (`y`), in other words, the effect that passes through the two mediators `z1` and `z1`

This (I think) is the product of the three individual mediator effects (`x` &rarr; `z1` &rarr; `z2` &rarr; `y`)

\[
  ACME = \beta_{xz_1} \times \beta_{z_1z_2} \times \beta_{z_2y}
  \]

```{r}
ACME = coef(fit2)['x'] * coef(fit3)['z1'] * coef(fit4)['z2']
ACME
```

Expected ACME is
```{r}
ACMEe = b_xz1 * b_z1z2 * b_z2y
ACMEe
```

##### ADE

`ADE` stands for average direct effects. This is the direct effect of the treatment on the outcome, and is `x` coefficient from step 3

```{r}
ADE = coef(fit4)['x']
ADE
```

##### Total effect

`Total Effect` stands for the total effect (direct + indirect). This is the same  as the coefficient from model 1 (`r round(coef(fit1)['x'], 2)`). We can also get it by simply adding the ACME and the ADE

```{r}
ACME + ADE
```

## Example 4: WNAI framework

This represents an indirect path from intervention to outcome via two dependent mediators (`x` &rarr; `z1` &rarr; `z2` &rarr; `y`), and a second mediation path (`x` &rarr; `z2` &rarr; `y`)

### Simulated data

- $x \sim N(5, 1)$
- $z_1 = \beta_{xz_1} \times x + e$
- $z_2 = \beta_{xz_2} \times x + \beta_{z_1z_2} z_1 + e$
- $y = \beta_{xy} \times x + \beta_{z_2y} \times z_2 + e$

```{r}
n <- 1000

b_xz1 = 10
b_xz2 = -3.5
b_z1z2 = -5
b_xy = 2.5
b_z2y = -0.5

x <- rnorm(n, 5, 1)
z1 <- x * b_xz1 + rnorm(n, 0, 5)
z2 <- x * b_xz2 + z1 * b_z1z2 + rnorm(n, 0, 5)
y <- x * b_xy + z2 * b_z2y + rnorm(n, 0, 5)
df <- data.frame(x, y, z1, z2)
```

```{r}
pairs(df)
```

### DAG

```{r}
model <- "
  z1 ~ b_xz1 * x
  z2 ~ b_xz2 * x + b_z1z2 * z1
  y ~ b_xy * x + b_z2y * z2
  ind_eff := (b_xz1 * b_z1z2 * b_z2y) + (b_xz2 * b_z2y)
  total := b_xy + ((b_xz1 * b_z1z2 * b_z2y) + (b_xz2 * b_z2y))
"
semfit <- lavaan::sem(model = model, data = df)
lavaanPlot(model = semfit)
```

### Mediation

#### Step 1: Total effect

```{r}
fit1 <- lm(y ~ x, df)
summary(fit1)
```

#### Step 2: Effect of IV on mediators

- `z1`

```{r}
fit2a <- lm(z1 ~ x, df)
summary(fit2a)
```

The expected value for this is `r b_xz1`, and we obtain a value of `r round(coef(fit2a)['x'], 2)`. 

- `z2`

```{r}
fit2b <- lm(z2 ~ x + z1, df)
summary(fit2b)
```

The expected value for the coefficient `x` is `r b_xz2`, and we obtain a value of `r round(coef(fit2b)['x'], 2)`. 

The expected value for the coefficient `z1` is `r b_z1z2`, and we obtain a value of `r round(coef(fit2b)['z1'], 2)`. 

#### Step 3: Effect of mediator (`z1`) on mediator (`z2`)

Not sure if we need this, but let's grab it anyway. Might need a control for `x`, which makes it the same as 2b

```{r}
fit3 <- lm(z2 ~ z1, df)
summary(fit3)
```

The expected value for this is `r b_z1z2`, and we obtain a value of `r round(coef(fit3)['z1'], 2)`. 

#### Step 4: Effect of mediator on DV

To get the effect of  `z2` on `y` we need to account for `x` as it is a fork or common cause (not shown here). WE DROP Z1 THOUGH - WHY?

```{r}
fit4 <- lm(y ~ x + z2, df)
summary(fit4)
```

The mediator (`z1` &rarr; `y`) is `r round(coef(fit4)['z1'], 2)`, which should be close to the known effect of `r b_z1y` (at least within CIs). 

The mediator (`z2` &rarr; `y`) is `r round(coef(fit4)['z2'], 2)`, which should be close to the known effect of `r b_z2y` (at least within CIs). 

The direct effect in this model (`x` &rarr; `y`) is `r round(coef(fit4)['x'], 2)`. As this is significant, this is *partial mediation*.

#### Step 5: Estimate mediation effects

```{r}
summary(semfit)
```

##### ACME 

`ACME` (average causal mediation effect). The indirect effect of the independent variable / treatment (`x`) on the outcome (`y`), in other words, the effect that passes through the two mediators `z1` and `z1`

With two possible mediation paths, we need to sum the effects so:

\[
ACME = \beta_{xz_1} \times \beta_{z_1z_2} \times \beta_{z_2y} + \beta_{x_z2} \times \beta_{z_2y}
\]

Note that if there was a path `z1` &rarr; `y`, we would add the product of those coefficients to the model. Now let's identify which models have these coefficients

- $\beta_{xz_1}$; `fit2a`: `r round(coef(fit2a)['x'], 2)`; Expected `r b_xz1`
- $\beta_{z_1z_2}$; `fit2b`: `r round(coef(fit2b)['z1'], 2)`; Expected `r b_z1z2`
- $\beta_{z_2y}$; `fit4`: `r round(coef(fit4)['z2'], 2)`; Expected `r b_z2y`
- $\beta_{xz_2}$; `fit2b`: `r round(coef(fit2b)['x'], 2)`; Expected `r b_xz2`

```{r}
ACME = coef(fit2a)['x'] * coef(fit2b)['z1'] * coef(fit4)['z2'] +
  coef(fit2b)['x'] * coef(fit4)['z2']
ACME
```

##### ADE

`ADE` stands for average direct effects. This is the direct effect of the treatment on the outcome, and is `x` coefficient from step 3

```{r}
ADE = coef(fit4)['x']
ADE
```

##### Total effect

`Total Effect` stands for the total effect (direct + indirect). This is the same  as the coefficient from model 1 (`r round(coef(fit1)['x'], 2)`). We can also get it by simply adding the ACME and the ADE

```{r}
ACME + ADE
```
