---
title: "WNAI Causal Diagram"
author: "Simon Brewer"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries and whatnot

```{r warning=FALSE, message=FALSE}
library(tidyverse)
# library(broom)
library(patchwork)
library(scales)
library(dagitty)
library(ggdag)
library(latex2exp)  # Easily convert LaTeX into arcane plotmath expressions
library(ggtext)     # Use markdown in ggplot labels

# Create a cleaner serifed theme to use throughout
theme_do_calc <- function() {
  theme_dag(base_family = "Times New Roman") +
    theme(plot.title = element_text(size = rel(1.5)),
          plot.subtitle = element_markdown())
}

# Make all geom_dag_text() layers use these settings automatically
update_geom_defaults(ggdag:::GeomDagText, list(family = "Times New Roman", 
                                               fontface = "bold",
                                               color = "black"))
```

## DAG 1

Model linking:

- climate (temperature and precipitation)
- society (density)
- animals (richness)
- plants (richness)


```{r}
wnai_dag <- dagify(density ~ temp + precip,
                    animal ~ temp + precip + density,
                    plant ~ temp + precip + density + animal,
                    coords = list(x = c(temp = 1, precip = 1, 
                                        density = 2, animal = 2,
                                        plant = 3),
                                  y = c(temp = 1, precip = 3, 
                                        density = 1, animal = 3,
                                        plant = 2)))

ggdag(wnai_dag) +
  theme_dag()
```

```{r}
impliedConditionalIndependencies(wnai_dag)
```

Adjustment set: shows what adjustment is necessary to obtain the causal link between input and outcome

```{r}
ggdag_adjustment_set(wnai_dag,
                    exposure = "density",
                    outcome = "plant") +
  theme_dag()
```

```{r}
ggdag_adjustment_set(wnai_dag,
                    exposure = "temp",
                    outcome = "plant") +
  theme_dag()
```

```{r}
ggdag_adjustment_set(wnai_dag,
                    exposure = "precip",
                    outcome = "plant") +
  theme_dag()
```

```{r}
ggdag_adjustment_set(wnai_dag,
                    exposure = "animal",
                    outcome = "plant") +
  theme_dag()
```

Path example (Density -> Plant)

```{r}
ggdag_paths(wnai_dag, from = "density", to = "plant",
            node_size = 12, text_size = 2.5)
```

```{r}
ggdag_adjustment_set(wnai_dag,
                    exposure = "density",
                    outcome = "plant") +
  theme_dag()
```


Colliders

```{r}
ggdag_collider(wnai_dag) +
  theme_dag()
```

## DAG 2 

Includes agriculture and language

```{r}
wnai_dag <- dagify(density ~ temp + precip,
                   agri  ~ density + temp + precip,
                    animal ~ temp + precip + density,
                    plant ~ temp + precip + density + animal + agri + language,
                    coords = list(x = c(temp = 1, precip = 1, 
                                        density = 2, animal = 2.5,
                                        agri = 3, language = 3.5,
                                        plant = 4),
                                  y = c(temp = 2, precip = 3, 
                                        density = 1, animal = 4,
                                        agri = 1, language = 4,
                                        plant = 2.5)))

ggdag(wnai_dag) +
  theme_dag()
```

```{r}
impliedConditionalIndependencies(wnai_dag)
```

Adjustment set: shows what adjustment is necessary to obtain the causal link between input and outcome

```{r}
ggdag_adjustment_set(wnai_dag,
                    exposure = "temp",
                    outcome = "plant") +
  theme_dag()
```

```{r}
ggdag_adjustment_set(wnai_dag,
                    exposure = "precip",
                    outcome = "plant") +
  theme_dag()
```

```{r}
ggdag_adjustment_set(wnai_dag,
                    exposure = "density",
                    outcome = "animal") +
  theme_dag()
```

```{r}
ggdag_adjustment_set(wnai_dag,
                    exposure = "density",
                    outcome = "plant") +
  theme_dag()
```

```{r}
ggdag_adjustment_set(wnai_dag,
                    exposure = "agri",
                    outcome = "plant") +
  theme_dag()
```

```{r}
ggdag_adjustment_set(wnai_dag,
                    exposure = "animal",
                    outcome = "plant") +
  theme_dag()
```

