---
title: "WNAI model"
author: "Simon Brewer"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
set.seed(1234)
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries and whatnot

```{r warning=FALSE, message=FALSE}
library(tidyverse)
# library(broom)
library(patchwork)
library(scales)
library(dagitty)
library(ggdag)
library(latex2exp)  # Easily convert LaTeX into arcane plotmath expressions
library(ggtext)     # Use markdown in ggplot labels

# Create a cleaner serifed theme to use throughout
theme_do_calc <- function() {
  theme_dag(base_family = "Times New Roman") +
    theme(plot.title = element_text(size = rel(1.5)),
          plot.subtitle = element_markdown())
}

# Make all geom_dag_text() layers use these settings automatically
update_geom_defaults(ggdag:::GeomDagText, list(family = "Times New Roman", 
                                               fontface = "bold",
                                               color = "black"))
```

## DAG

```{r}
wnai_dag <- dagify(density ~ temp + precip,
                   agri  ~ density + temp + precip,
                   plant ~ temp + precip + density + agri,
                   animal ~ temp + precip + density + plant + language,
                   coords = list(x = c(temp = 1, precip = 1, 
                                       density = 2, plant = 2.5,
                                       agri = 3, language = 3.5,
                                       animal = 4),
                                 y = c(temp = 2, precip = 3, 
                                       density = 1, plant = 4,
                                       agri = 1, language = 4,
                                       animal = 2.5)))

ggdag(wnai_dag) +
  theme_dag()
```

## Simulated data

```{r}
n = 10000
precip = rlnorm(n, log(20), log(2))
precip = rnorm(n, 200, 20)
temp = rnorm(n, 0, 1)
lang = sample(c(0,1,2,3), n, replace = TRUE)
```

```{r}
density = 10 * precip + 200 * temp + rnorm(n, 0, 250)
agri = 2.5 * precip + 10 * temp + .5 * density + rnorm(n, 0, 150)

b_da = -0.5 # Coefficient for animal ~ density
plant = 1000 + (2 * precip) + (75 * temp) + b_da * density - 0.05 * agri  + rnorm(n, 0, 50)

b_dp = -0.5 # Coefficient for plant ~ density
## Plants with no animal link (test for spurious correlations)
animal = 1000 + 4 * precip + 10 * temp + b_dp * density+ 1 * lang + rnorm(n, 0, 50)

## Plants with animal link
animal = 1000 + 4 * precip + 10 * temp + b_dp * density + 0.25 * plant + 1 * lang + rnorm(n, 0, 50)

```

```{r}
df = data.frame(temp, precip, lang, 
                density, agri, plant, animal)
```

Scatter plots

```{r}
pairs(df)
```

Quick vif check

```{r}
fit <- lm(animal ~ density + precip + temp + plant + agri + lang, df)
car::vif(fit)
```

## Model 1: Plants

Causal link to be tested: `plants ~ density`. 'Real' value is `r round(b_dp, 2)`.

Let's check the adjustment set for this. There are two forks (via `precip` and `temp`), which create backdoor paths that need to be controlled for. In addition, `agri` forms a chain through to plant richness and will need to be controlled for

```{r}
ggdag_adjustment_set(wnai_dag, 
                     exposure = "density",
                     outcome = "plant",
                     shadow = TRUE) +
  theme_dag()
```

There is also a potential collider (`animal`), which should be avoided. As long as we leave this alone, we don't introduce any correlation

```{r}
ggdag_collider(wnai_dag) +
  theme_dag()
```


### 1a: Naive model

Only the direct path from `density` to `plant`

```{r}
fit <- lm(plant ~ density, df)
summary(fit)
```

`density` coefficient: `r round(coef(fit)['density'], 2)`

### 1b

Include controls for backdoor confounders *and* collider (`animal`)

```{r}
fit <- lm(plant ~ density + precip + temp + agri + animal, df)
summary(fit)
```

`density` coefficient: `r round(coef(fit)['density'], 2)`

### 1c

Final model (ignore colliders)

```{r}
fit <- lm(plant ~ density + precip + temp + agri, df)
summary(fit)
```

`density` coefficient: `r round(coef(fit)['density'], 2)`

### 1d

Example adding independent variable (it's not correlated), but as it's independent, it does not bias the estimate

```{r}
fit <- lm(plant ~ density + precip + temp + agri + lang, df)
summary(fit)
```

`density` coefficient: `r round(coef(fit)['density'], 2)`

### tl;dr

Testing the causal link between plant richness and population density requires

- Closing the backdoor paths (precip and temp)
- Adding `agri` as a confounder
- Not adding `animal` as a collidor

## Model 2: Animal

Causal link to be tested: `animal ~ density`. 'Real' value is `r round(b_da, 2)`.

Let's check the adjustment set for this. There are the same two forks (via `precip` and `temp`), which create backdoor paths that need to be controlled for, There are *two* chains now

- `density` &rarr; `plant` &rarr; `animal` 
- `density` &rarr; `agri` &rarr; `animal` 

And we'll need to control for these two intermediate variables

```{r}
ggdag_adjustment_set(wnai_dag, 
                     exposure = "density",
                     outcome = "animal",
                     shadow = TRUE) +
  theme_dag()
```

While there are the same set of colliders, none are children of `animal` so we don't need to worry about adjusting for these

```{r}
ggdag_collider(wnai_dag) +
  theme_dag()
```

### 2a: Naive model

Only the direct path from `density` to `animal`

```{r}
fit <- lm(animal ~ density, df)
summary(fit)
```

`density` coefficient: `r round(coef(fit)['density'], 2)`

### 2b

Include controls for backdoor confounders

```{r}
fit <- lm(animal ~ density + precip + temp, df)
summary(fit)
```

`density` coefficient: `r round(coef(fit)['density'], 2)`

### 2c

Final model (all possible confounders (except agriculture as there is no path?)

```{r}
fit <- lm(animal ~ density + precip + temp + plant + lang + agri, df)
summary(fit)
```

`density` coefficient: `r round(coef(fit)['density'], 2)`

### tl;dr

Testing the causal link between animal richness and population density requires

- Closing the backdoor paths (precip and temp)
- Adding `plant` as a confounder
- Not including (or optionally including `agri`). Although this is connected to `animal` (see d-sep diagram), this path is closed by controlling for `plant`

```{r}
ggdag_dseparated(wnai_dag, from = "agri", to = "animal")
```

To illustrate, here are the open paths between `density` and `animal` without a control on `plant` (but with controls on the backdoor paths). Note that the path through `agri` remains open. 

```{r}
ggdag_paths(wnai_dag, from = "density", to = "animal", 
            adjust_for = c("temp", "precip"))
```

Here is the same with a control added for `plant`, which only leaves the direct path:

```{r}
ggdag_paths(wnai_dag, from = "density", to = "animal", 
            adjust_for = c("temp", "precip", "plant"))
```