---
title: "WNAI model"
author: "Simon Brewer"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
set.seed(1234)
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries and whatnot

```{r warning=FALSE, message=FALSE}
library(lavaan)
library(lavaanPlot)
library(dagitty)
library(ggdag)
```

## Full DAG

```{r}
wnai_dag <- dagify(density ~ temp + precip,
                   agri  ~ density + temp + precip,
                   plant ~ temp + precip + density + agri,
                   animal ~ temp + precip + density + plant + agri,
                   coords = list(x = c(temp = 1, precip = 1, 
                                       density = 2, plant = 2.5,
                                       agri = 3,
                                       animal = 4),
                                 y = c(temp = 2, precip = 3, 
                                       density = 1, plant = 4,
                                       agri = 1,
                                       animal = 2.5)))

ggdag(wnai_dag) +
  theme_dag()
```

## Data

```{r}
dat <- read.csv("~/Dropbox/DB Docs/codding/causal_paper/data/WNAI-DPLACE_Causal_060223_v2.csv")
```

```{r}
df <- data.frame(precip = dat$MonthlyMeanPrecip,
                 temp = dat$AnnualMeanTemperature,
                 density = dat$WNAI288_PopDensityAve,
                 agri = dat$WNAI189_Ag01,
                 animal = dat$LandMammal_Richness,
                 plant = dat$Plant_Richness
)
```

```{r}
skimr::skim(df)
```

```{r}
fit <- lm(animal ~ density + precip + temp + plant + agri, df)
car::vif(fit)
```

## Model 1: Plants

Causal link to be tested: `plants ~ agri`. 

Let's check the adjustment set for this. There are two forks (via `precip` and `temp`), which create backdoor paths that need to be controlled for. In addition, `agri` forms a chain through to plant richness and will need to be controlled for

```{r}
ggdag_adjustment_set(wnai_dag, 
                     exposure = "agri",
                     outcome = "plant",
                     shadow = TRUE) +
  theme_dag()
```

### DAG

model <- "
  z ~ b_xz * x
  y ~ b_xy * x + b_zy * z
  ind_eff := b_xz * b_zy
  total := b_xy + (b_xz * b_zy)
"

```{r}
df2 <- as.data.frame(scale(df))
model <- "
  density ~ b_td * temp + b_pd * precip
  agri ~ b_tag * temp + b_pag * precip + b_dag * density
  plant ~ b_tpl * temp + b_ppl * precip + b_dpl * density + b_agpl * agri
"
semfit <- lavaan::sem(model = model, data = df2)
lavaanPlot(model = semfit)
```

### Mediation

#### Total Effect 

Note that this model has *no* mediation. So we simply need to control for the confounders

```{r}
fit1 <- lm(plant ~ agri + temp + precip + density, df2)
summary(fit1)
```

```{r}
summary(semfit)
```


## Model 2: Animals

Causal link to be tested: `animal ~ agri`. 

Let's check the adjustment set for this. There are three forks (via `density`, `precip` and `temp`), which create backdoor paths that need to be controlled for. In addition, there is a chain through `plant` which will cause an indirect effect

```{r}
ggdag_adjustment_set(wnai_dag, 
                     exposure = "agri",
                     outcome = "animal",
                     shadow = TRUE) +
  theme_dag()
```

### DAG

```{r}
df2 <- as.data.frame(scale(df))
model <- "
  density ~ b_td * temp + b_pd * precip
  agri ~ b_tag * temp + b_pag * precip + b_dag * density
  plant ~ b_tpl * temp + b_ppl * precip + b_dpl * density + b_agpl * agri
  animal ~ b_tan * temp + b_pan * precip + b_dan * density + b_plan * plant + b_agan * agri
  ind_eff := b_agpl * b_plan
  total := b_agan + ind_eff
"
semfit <- lavaan::sem(model = model, data = df2)
lavaanPlot(model = semfit)
```

### Mediation

#### Step 1: Total effect

We do not control for the chain here

```{r}
fit1 <- lm(animal ~ agri + density + temp + precip, df2)
summary(fit1)
```

The total effect is given by the product of the coefficients of the indirect path plus the coefficient of the direct path: $TE = \beta_{xz} \times \beta_{zy} + \beta_{xy}$. From this model we get a value of `r round(coef(fit1)['agri'], 2)`. 

#### Step 2: Effect of IV on mediator

```{r}
ggdag_adjustment_set(wnai_dag, 
                     exposure = "agri",
                     outcome = "plant",
                     shadow = TRUE) +
  theme_dag()

```

Again, we need to control for the backdoor paths through `density`, `temp` and `precip`

```{r}
fit2 <- lm(plant ~ agri + density + temp + precip, df2)
summary(fit2)
```

We obtain a value of `r round(coef(fit2)['agri'], 2)`. 

#### Step 3: Effect of mediator on DV

To get the effect of `plant` on `animal` we need to account for `agri` as it is a fork or common cause, as well as the usual backdoor paths

```{r}
ggdag_adjustment_set(wnai_dag, 
                     exposure = "plant",
                     outcome = "animal",
                     shadow = TRUE) +
  theme_dag()

```

```{r}
fit3 <- lm(animal ~ plant + agri + density + temp + precip, df2)
summary(fit3)
```

The mediator (`z` &rarr; `y`) is `r round(coef(fit3)['plant'], 2)`.

#### Step 4: Estimate mediation effects

```{r}
summary(semfit)
```

```{r}
lavaanPlot(model = semfit, coefs = TRUE, stars = "regress")
```

##### ACME

`ACME` (average causal mediation effect). This is the effect of the treatment on the mediator (`r round(coef(fit2)['agri'], 2)`) from step 2 times the mediator's effect on the outcome from step 3 (`r round(coef(fit3)['plant'], 2)`)

```{r}
ACME = coef(fit2)['agri'] * coef(fit3)['plant']
ACME
```

##### ADE

`ADE` stands for average direct effects. This is the direct effect of the treatment on the outcome, and is the same as the `agri` coefficient from step 3

```{r}
ADE = coef(fit3)['agri']
ADE
```

##### Total Effect

`Total Effect` stands for the total effect (direct + indirect). This is the same  as the `density` coefficient from model 1 (`r round(coef(fit1)['density'], 2)`). 

Note we can also get this as the sum of ACME and ADE

```{r}
TE = ACME + ADE
TE
```

##### Proportion Mediated 

`Prop. Mediated` describes the proportion of the effect of the IV on the DV that goes through the mediator. This is calculated by dividing the ACME by the total effect

```{r}
ACME / TE
```

