---
title: "WNAI model"
author: "Simon Brewer"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
set.seed(42)
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries and whatnot

```{r warning=FALSE, message=FALSE}
library(tidyverse)
# library(broom)
library(patchwork)
library(scales)
library(dagitty)
library(ggdag)
library(latex2exp)  # Easily convert LaTeX into arcane plotmath expressions
library(ggtext)     # Use markdown in ggplot labels

# Create a cleaner serifed theme to use throughout
theme_do_calc <- function() {
  theme_dag(base_family = "Times New Roman") +
    theme(plot.title = element_text(size = rel(1.5)),
          plot.subtitle = element_markdown())
}

# Make all geom_dag_text() layers use these settings automatically
update_geom_defaults(ggdag:::GeomDagText, list(family = "Times New Roman", 
                                               fontface = "bold",
                                               color = "black"))
```

## DAG

```{r}
wnai_dag <- dagify(density ~ temp + precip,
                   agri  ~ density + temp + precip,
                    animal ~ temp + precip + density,
                    plant ~ temp + precip + density + animal + agri + language,
                    coords = list(x = c(temp = 1, precip = 1, 
                                        density = 2, animal = 2.5,
                                        agri = 3, language = 3.5,
                                        plant = 4),
                                  y = c(temp = 2, precip = 3, 
                                        density = 1, animal = 4,
                                        agri = 1, language = 4,
                                        plant = 2.5)))

ggdag(wnai_dag) +
  theme_dag()
```

## Simulated data

- `precip`: $lognorm(20, 5)$
- `temp`: $N(0, 5)$
- `lang`: $[0,3]$
- `density`: $10 \times precip + 20 \times temp + rnorm(0, 10)$
- `animal`: $10 \times precip + 30 \times temp - 5 \times density + rnorm(0, 10)$
- `agri`: $25 \times precip + 10 \times temp + 10 \times density + rnorm(0, 10)$
- `plant`: $4 \times precip + 2 \times temp - 2 \times density - 3 \times agri + 2 \times lang + rnorm(0, 10)$

```{r}
n = 1000
precip = rlnorm(n, log(20), log(2))
precip = rnorm(n, 200, 20)
temp = rnorm(n, 0, 1)
lang = sample(c(0,1,2,3), n, replace = TRUE)
```

```{r}
density = 10 * precip + 200 * temp + rnorm(n, 0, 5)
animal = 1000 + (2 * precip) + (50 * temp) - 0.5 * density + rnorm(n, 0, 5)
agri = 25 * precip + 10 * temp + 10 * density + rnorm(n, 0, 50)

plant = 1e5 + 4 * precip + 10 * temp - 0.1 * density - 0.1 * agri + 1e4 * lang + rnorm(n, 0, 50)

plant = 1e5 + 4 * precip + 10 * temp - 0.1 * density - 0.1 * agri + 1 * lang + rnorm(n, 0, 50)
```

```{r}
plot(precip, density)
plot(temp, density)
```

```{r}
plot(density, animal)
```

```{r}
plot(density, agri)
```

```{r}
plot(plant, precip)
plot(plant, temp)
plot(plant, density)
plot(plant, agri)
plot(plant, animal)
plot(plant, lang)
```

```{r}
df = data.frame(temp, precip, lang, 
                density, animal, agri, plant)
```

## Model 1: Animal

Causal link to be tested: `animal ~ density`. Let's check the adjustment set for this. There are two forks (via `precip` and `temp`), which create backdoor paths that need to be controlled for

```{r}
ggdag_adjustment_set(wnai_dag, 
                     exposure = "density",
                     outcome = "animal",
                     shadow = TRUE) +
  theme_dag()
```

There are also two potential colliders (`plant` and `agri`), which should be avoided. As long as we leave these alone, we don't introduce any correlation

```{r}
ggdag_collider(wnai_dag) +
  theme_dag()
```


### Naive model

Only the direct path from `density` to `animal`

```{r}
fit <- lm(animal ~ density, df)
summary(fit)
```

`density` coefficient: `r round(coef(fit)['density'], 2)`

### Model 2

Include controls for backdoor confounders *and* collider (`plant`)

```{r}
fit <- lm(animal ~ density + precip + temp + plant + agri, df)
summary(fit)
```

`density` coefficient: `r round(coef(fit)['density'], 2)`

### Model 3

Final model (ignore colliders)

```{r}
fit <- lm(animal ~ density + precip + temp, df)
summary(fit)
```

`density` coefficient: `r round(coef(fit)['density'], 2)`

### Model 4

Example adding independent variable (it's not correlated), but as it's independent, it does not bias the estimate

```{r}
fit <- lm(animal ~ density + precip + temp + lang, df)
summary(fit)
```

`density` coefficient: `r round(coef(fit)['density'], 2)`

## Model 1: Animal

Causal link to be tested: `plant ~ density`. Let's check the adjustment set for this. There are the same two forks (via `precip` and `temp`), which create backdoor paths that need to be controlled for

```{r}
ggdag_adjustment_set(wnai_dag, 
                     exposure = "density",
                     outcome = "plant",
                     shadow = TRUE) +
  theme_dag()
```

While there are the same set of colliders, none are children of `plant` so we don't need to worry about adjusting for these

```{r}
ggdag_collider(wnai_dag) +
  theme_dag()
```

### Naive model

Only the direct path from `density` to `plant`

```{r}
fit <- lm(plant ~ density, df)
summary(fit)
```

`density` coefficient: `r round(coef(fit)['density'], 2)`

### Model 2

Include controls for backdoor confounders

```{r}
fit <- lm(plant ~ density + precip + temp, df)
summary(fit)
```

`density` coefficient: `r round(coef(fit)['density'], 2)`

### Model 3

Final model (ignore colliders)

```{r}
fit <- lm(plant ~ density + precip + temp + animal + agri + lang, df)
summary(fit)
```

`density` coefficient: `r round(coef(fit)['density'], 2)`

### Model 4

Example adding independent variable (it's not correlated), but as it's independent, it does not bias the estimate

```{r}
fit <- lm(animal ~ density + precip + temp + lang, df)
summary(fit)
```

`density` coefficient: `r round(coef(fit)['density'], 2)`

