---
title: "WNAI model"
author: "Simon Brewer"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
set.seed(42)
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries and whatnot

```{r warning=FALSE, message=FALSE}
library(dagitty)
library(ggdag)
library(lme4)
library(stargazer)
```

## Full DAG

```{r}
wnai_dag <- dagify(density ~ temp + precip,
                   agri  ~ density + temp + precip,
                   plant ~ temp + precip + density + agri,
                   animal ~ temp + precip + density + plant + agri,
                   coords = list(x = c(temp = 1, precip = 1, 
                                       density = 2, plant = 2.5,
                                       agri = 3,
                                       animal = 4),
                                 y = c(temp = 2, precip = 3, 
                                       density = 1, plant = 4,
                                       agri = 1,
                                       animal = 2.5)))

ggdag(wnai_dag) +
  theme_dag()
```

## Data

```{r}
dat <- read.csv("~/Dropbox/DB Docs/codding/causal_paper/data/WNAI-DPLACE_Causal_060223_v2.csv")
```

```{r}
df <- data.frame(precip = scale(dat$MonthlyMeanPrecip),
                 temp = scale(dat$AnnualMeanTemperature),
                 density = scale(dat$WNAI288_PopDensityAve),
                 agri = dat$WNAI189_Ag01,
                 animal = dat$LandMammal_Richness,
                 plant = dat$Plant_Richness,
                 lang = dat$Language_family
)
```

```{r}
skimr::skim(df)
```

```{r}
fit <- lm(animal ~ density + precip + temp + plant + agri, df)
car::vif(fit)
```

## Model 1: Plants

```{r}
plant_dag <- dagify(density ~ temp + precip,
                   agri  ~ density + temp + precip,
                   plant ~ temp + precip + density + agri,
                   coords = list(x = c(temp = 1, precip = 1, 
                                       density = 2, plant = 2.5,
                                       agri = 3),
                                 y = c(temp = 2, precip = 3, 
                                       density = 1, plant = 4,
                                       agri = 1)))

ggdag(plant_dag) +
  theme_dag()
```

Causal link to be tested: `plants ~ agri`. 

Let's check the adjustment set for this. There are two forks (via `precip` and `temp`), which create backdoor paths that need to be controlled for. In addition, `agri` forms a chain through to plant richness and will need to be controlled for

```{r}
ggdag_adjustment_set(plant_dag, 
                     exposure = "agri",
                     outcome = "plant",
                     shadow = TRUE) +
  theme_dag()
```

### Mediation

#### Total Effect 

Note that this model has *no* mediation. So we simply need to control for the confounders

```{r}
fit1 <- lmer(plant ~ agri + density + temp + precip + (1 | lang), 
             data = df)
summary(fit1)
```

Total effect is `r round(fixef(fit1)['agri'], 2)`

```{r results = 'asis'}
stargazer(fit1, type = 'html')
```

#### Naive model for comparison

This is asking the question - what would happen if we ignore the confounders

```{r}
fit0 <- lmer(plant ~ agri + (1 | lang), 
             data = df)
summary(fit0)
```

Total effect is `r round(fixef(fit0)['agri'], 2)`

```{r results = 'asis'}
stargazer(fit0, type = 'html')
```

## Model 2: Animals

```{r}
animal_dag <- dagify(density ~ temp + precip,
                   agri  ~ density + temp + precip,
                   plant ~ temp + precip + density + agri,
                   animal ~ temp + precip + density + plant + agri,
                   coords = list(x = c(temp = 1, precip = 1, 
                                       density = 2, plant = 2.5,
                                       agri = 3,
                                       animal = 4),
                                 y = c(temp = 2, precip = 3, 
                                       density = 1, plant = 4,
                                       agri = 1,
                                       animal = 2.5)))

ggdag(animal_dag) +
  theme_dag()
```

Causal link to be tested: `animal ~ agri`. 

Let's check the adjustment set for this. There are three forks (via `density`, `precip` and `temp`), which create backdoor paths that need to be controlled for. In addition, there is a chain through `plant` which will cause an indirect effect

```{r}
ggdag_adjustment_set(animal_dag, 
                     exposure = "agri",
                     outcome = "animal",
                     shadow = TRUE) +
  theme_dag()
```

### Mediation

#### Step 1: Total effect

We do not control for the chain here

```{r}
fit1 <- lmer(animal ~ agri + density + temp + precip + (1 | lang), 
             data = df)
summary(fit1)
```

The total effect is given by the product of the coefficients of the indirect path plus the coefficient of the direct path: $TE = \beta_{xz} \times \beta_{zy} + \beta_{xy}$. From this model we get a value of `r round(fixef(fit1)['agri'], 2)`. 
```{r results = 'asis'}
stargazer(fit1, type = 'html')
```

#### Step 2: Effect of IV on mediator

```{r}
ggdag_adjustment_set(animal_dag, 
                     exposure = "agri",
                     outcome = "plant",
                     shadow = TRUE) +
  theme_dag()

```

Again, we need to control for the backdoor paths through `density`, `temp` and `precip`

```{r}
fit2 <- lmer(plant ~ agri + density + temp + precip + (1 | lang), 
             data = df)
summary(fit2)
```

We obtain a value of `r round(fixef(fit2)['agri'], 2)`. 

```{r results = 'asis'}
stargazer(fit2, type = 'html')
```

#### Step 3: Effect of mediator on DV

To get the effect of `plant` on `animal` we need to account for `agri` as it is a fork or common cause, as well as the usual backdoor paths

```{r}
ggdag_adjustment_set(animal_dag, 
                     exposure = "plant",
                     outcome = "animal",
                     shadow = TRUE) +
  theme_dag()

```

```{r}
fit3 <- lmer(animal ~ plant + agri + density + temp + precip + (1 | lang), 
             data = df)
summary(fit3)
```

The mediator (`z` &rarr; `y`) is `r round(fixef(fit3)['plant'], 2)`.

```{r results = 'asis'}
stargazer(fit3, type = 'html')
```

#### Step 4: Estimate mediation effects

##### ACME

`ACME` (average causal mediation effect). This is the effect of the treatment on the mediator (`r round(fixef(fit2)['agri'], 2)`) from step 2 times the mediator's effect on the outcome from step 3 (`r round(fixef(fit3)['plant'], 2)`)

```{r}
ACME = fixef(fit2)['agri'] * fixef(fit3)['plant']
ACME
```

##### ADE

`ADE` stands for average direct effects. This is the direct effect of the treatment on the outcome, and is the same as the `agri` coefficient from step 3

```{r}
ADE = fixef(fit3)['agri']
ADE
```

##### Total Effect

`Total Effect` stands for the total effect (direct + indirect). This is the same  as the `density` coefficient from model 1 (`r round(fixef(fit1)['agri'], 2)`). 

Note we can also get this as the sum of ACME and ADE

```{r}
TE = ACME + ADE
TE
```

##### Proportion Mediated 

`Prop. Mediated` describes the proportion of the effect of the IV on the DV that goes through the mediator. This is calculated by dividing the ACME by the total effect

```{r}
ACME / TE
```

### Mediation check

```{r}
res = mediation::mediate(model.m = fit2, 
                         model.y = fit3, 
                         treat = 'agri',
                         mediator = 'plant', boot = FALSE)
summary(res)
```

