---
title: "WNAI model"
author: "Simon Brewer"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
set.seed(1234)
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries and whatnot

```{r warning=FALSE, message=FALSE}
library(tidyverse)
# library(broom)
library(patchwork)
library(scales)
library(dagitty)
library(ggdag)
library(latex2exp)  # Easily convert LaTeX into arcane plotmath expressions
library(ggtext)     # Use markdown in ggplot labels
library(mediation)  # Mediation analysis

# Create a cleaner serifed theme to use throughout
theme_do_calc <- function() {
  theme_dag(base_family = "Times New Roman") +
    theme(plot.title = element_text(size = rel(1.5)),
          plot.subtitle = element_markdown())
}

# Make all geom_dag_text() layers use these settings automatically
update_geom_defaults(ggdag:::GeomDagText, list(family = "Times New Roman", 
                                               fontface = "bold",
                                               color = "black"))
```

## DAG

```{r}
wnai_dag <- dagify(density ~ temp + precip,
                   agri  ~ density + temp + precip,
                   plant ~ temp + precip + density + agri,
                   animal ~ temp + precip + density + plant,
                   coords = list(x = c(temp = 1, precip = 1, 
                                       density = 2, plant = 2.5,
                                       agri = 3, animal = 4),
                                 y = c(temp = 2, precip = 3, 
                                       density = 1, plant = 4,
                                       agri = 1, animal = 2.5)))

ggdag(wnai_dag) +
  theme_dag()
```

## Simulated data

```{r}
n = 1000
precip = rlnorm(n, log(20), log(2))
precip = rnorm(n, 200, 20)
temp = rnorm(n, 0, 1)
lang = sample(c(0,1,2,3), n, replace = TRUE)
```

```{r}
density = 10 * precip + 200 * temp + rnorm(n, 0, 250)
agri = 2.5 * precip + 10 * temp + .5 * density + rnorm(n, 0, 150)

b_da = -0.5 # Coefficient for animal ~ density
plant = 1000 + (2 * precip) + (75 * temp) + b_da * density - 0.05 * agri  + rnorm(n, 0, 50)

b_dp = -0.5 # Coefficient for plant ~ density
## Plants with no animal link (test for spurious correlations)
animal = 1000 + 4 * precip + 10 * temp + b_dp * density + rnorm(n, 0, 50)

## Plants with animal link
animal = 1000 + 4 * precip + 10 * temp + b_dp * density + 0.25 * plant + rnorm(n, 0, 50)

```

```{r}
df = data.frame(temp, precip, lang, 
                density, agri, plant, animal)
```

Scatter plots

```{r}
pairs(df)
```

Quick vif check

```{r}
fit <- lm(animal ~ density + precip + temp + plant + agri, df)
car::vif(fit)
```

## Model 1: Plants

Causal link to be tested: `plants ~ density`. 'Real' value is `r round(b_dp, 2)`.

Let's check the adjustment set for this. There are two forks (via `precip` and `temp`), which create backdoor paths that need to be controlled for. In addition, `agri` forms a chain through to plant richness and will need to be controlled for

```{r}
ggdag_adjustment_set(wnai_dag, 
                     exposure = "density",
                     outcome = "plant",
                     shadow = TRUE) +
  theme_dag()
```

There is also a potential collider (`animal`), which should be avoided. As long as we leave this alone, we don't introduce any correlation

```{r}
ggdag_collider(wnai_dag) +
  theme_dag()
```


### 1a: Naive model

Only the direct path from `density` to `plant`

```{r}
fit <- lm(plant ~ density, df)
summary(fit)
```

`density` coefficient: `r round(coef(fit)['density'], 2)`

### 1b: Causal model

Include controls for backdoor confounders. This should estimate the full causal effect of `density` on `plant`

```{r}
fit <- lm(plant ~ density + precip + temp, df)
summary(fit)
```

`density` coefficient: `r round(coef(fit)['density'], 2)`

### 1c: Mediation

Mediation path is through agriculture

```{r}
ggdag_paths(wnai_dag, from = 'density', to = 'plant', 
            adjust_for = c('temp', 'precip')) +
  theme_dag()
```

#### Step 1c-1: Total effect

This is the equivalent to the total causal effect. As this is a chain (`density` &rarr; `agri` &rarr; `plant`), we do *not* need to control for the mediator. We need to keep the controls for the backdoor paths through `temp` and `precip`. Ignoring the backdoor controls this is:

\[
\mbox{plant} \sim \beta_{te} \times \mbox{density} 
\]

```{r}
fit1 = lm(plant ~ density + temp + precip, df)
summary(fit1)
```

The total effect is `r round(coef(fit1)['density'], 2)`. 

#### Step 1c-2: The effect of the IV on the mediator

Next we need to quantify the effect of the independent variable (`density`) on the mediator (`agri`). Again, we need to control for the backdoor paths through `precip` and `temp`

\[
\mbox{agri} \sim \beta_{te} \times \mbox{density} 
\]

```{r}
ggdag_adjustment_set(wnai_dag, 
                     exposure = 'density', 
                     outcome = 'agri') +
  theme_dag()
```

```{r}
fit2 = lm(agri ~ density + temp + precip, df)
summary(fit2)
```

And we get a result of `r round(coef(fit2)['density'],2)`.

#### Step 1c-3: The effect of the mediator on the dependent variable

The third step confirms that the mediator affects the dependent variable while *controlling* for the independent variable (and the backdoors!). This means, that for a mediation to take place, the mediator must explain more or other parts of the variance in the dependent variable than the independent variable.

```{r}
fit3 = lm(plant ~ density + agri + temp + precip, df)
summary(fit3)
```

#### Step 1c-4: Causal Mediation Analysis

```{r}
res = mediation::mediate(model.m = fit2, 
                         model.y = fit3, 
                         treat = 'density',
                         mediator = 'agri',
                         boot = TRUE)
```

```{r}
summary(res)
```


- `ACME` stands for average causal mediation effects. This is the indirect effect of the IV (sepal length) on the DV (likelihood of pollination) that goes through the mediator (attractiveness to bee). Note that it estimated this effect to be `r round(res$d0, 3)`. This is the effect of the treatment on the mediator (`r round(coef(fit2)['density'], 2)`) from step 2 times the mediator's effect on the outcome from step 3 (`r round(coef(fit3)['agri'], 2)`)
- `ADE` stands for average direct effects. This is the direct effect of the treatment on the outcome. This is the same as the `sl` coefficient from step 3
- `Total Effect` stands for the total effect (direct + indirect), which was calculated in step #1. We can also get it by simply adding the ACME (`r round(res$d0, 3)`) and the ADE (`r round(res$z0, 3)`) to receive the total effect of `r round(res$tau.coef, 3)`. 
- `Prop. Mediated` describes the proportion of the effect of the IV on the DV that goes through the mediator. It’s calculated by dividing the ACME (`r round(res$d0, 3)`) by the total effect (`r round(res$tau.coef, 3)`) to receive a value of roughly `r round(res$n0, 3)`. This supports the results above that indicate there is no direct effect. 

```{r}
plot(res)
```

## Model 2: Animals

Causal link to be tested: `animal ~ density`. 'Real' value is `r round(b_da, 2)`.

Let's check the adjustment set for this. There are two forks (via `precip` and `temp`), which create backdoor paths that need to be controlled for. In addition, `agri` forms a chain through to plant richness and will need to be controlled for

```{r}
ggdag_adjustment_set(wnai_dag, 
                     exposure = "density",
                     outcome = "animal",
                     shadow = TRUE) +
  theme_dag()
```

There is also a potential collider (`animal`), which should be avoided. As long as we leave this alone, we don't introduce any correlation

```{r}
ggdag_collider(wnai_dag) +
  theme_dag()
```


### 2a: Naive model

Only the direct path from `density` to `animal`

```{r}
fit <- lm(animal ~ density, df)
summary(fit)
```

`density` coefficient: `r round(coef(fit)['density'], 2)`

### 2b: Causal model

Include controls for backdoor confounders. This should estimate the full causal effect of `density` on `animal`. Note that we should adjust for plant (collider) or `agri` (chain / overcontrol)

```{r}
fit <- lm(animal ~ density + precip + temp, df)
summary(fit)
```

`density` coefficient: `r round(coef(fit)['density'], 2)`

### 2c: Mediation

Mediation path is through `agri` *and* `plant`

```{r}
ggdag_paths(wnai_dag, from = 'density', to = 'animal', 
            adjust_for = c('temp', 'precip')) +
  theme_dag()
```

#### Step 2c-1: Total effect

This is the equivalent to the total causal effect. As this is a chain (`density` &rarr; `agri` &rarr; `plant`), we do *not* need to control for the mediator. We need to keep the controls for the backdoor paths through `temp` and `precip`

```{r}
fit1 = lm(animal ~ density + temp + precip, df)
summary(fit1)
```

The total effect is `r round(coef(fit1)['density'], 2)`. 

#### Step 2c-2: The effect of the IV on the mediator

Next we need to quantify the effect of the independent variable (`density`) on the two mediators. Again, we need to control for any backdoor paths.

- Agriculture

```{r}
ggdag_adjustment_set(wnai_dag, 
                     exposure = 'density', 
                     outcome = 'agri') +
  theme_dag()
```

```{r}
fit2a = lm(agri ~ density + temp + precip, df)
summary(fit2a)
```

And we get a result of `r round(coef(fit2a)['density'],2)`.

- Plant

```{r}
ggdag_adjustment_set(wnai_dag, 
                     exposure = 'density', 
                     outcome = 'plant') +
  theme_dag()
```

```{r}
fit2b = lm(plant ~ density + temp + precip, df)
summary(fit2b)
```

And we get a result of `r round(coef(fit2b)['density'],2)`.


